---
title: "CMPINF 2130 Summer 2021 - Week 07"
author: "Dr. Joseph P. Yurko"
date: "6/28/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load tidyverse

```{r, load_tidyverse_pkg}
library(tidyverse)
```

## Boxplot review

Work with the synthetic from a standard normal.  

```{r, make_fake_data}
set.seed(2130)

x <- rnorm(n = 5000)

xdf <- tibble::tibble(x = x)

xdf %>% glimpse()
```

Make a boxplot.  

```{r, viz_fake_data_box}
xdf %>% 
  ggplot(mapping = aes(y = x)) +
  geom_boxplot() +
  scale_x_discrete() +
  theme_bw()
```

Calculate the whisker length above and below the box.  

```{r, fake_data_whiskers}
xdf %>% 
  summarise(q25 = quantile(x, 0.25),
            q50 = median(x),
            q75 = quantile(x, 0.75)) %>% 
  mutate(whisker_low = q25 - 1.5 * (q75 - q25),
         whisker_high = q75 + 1.5 * (q75 - q25))
```

Let's visualize these components with `geom_linerange()`.  

```{r, viz_manual_boxplot}
xdf %>% 
  summarise(q25 = quantile(x, 0.25),
            q50 = median(x),
            q75 = quantile(x, 0.75)) %>% 
  mutate(whisker_low = q25 - 1.5 * (q75 - q25),
         whisker_high = q75 + 1.5 * (q75 - q25)) %>% 
  mutate(var_name = 'x') %>% 
  ggplot(mapping = aes(x = var_name)) +
  geom_linerange(mapping = aes(ymin = whisker_low,
                               ymax = whisker_high),
                 color = 'black', size = 1.1) +
  geom_linerange(mapping = aes(ymin = q25, ymax = q75),
                 color = 'grey', size = 9) +
  geom_point(mapping = aes(y = q50),
             shape = 3, color = 'black', size = 5.5) +
  labs(y = "variable value") +
  theme_bw()
```

Create variables for the whisker low and high points.  

```{r, make_whisker_variables}
gauss_whisker_low <- xdf %>% 
  summarise(q25 = quantile(x, 0.25),
            q50 = median(x),
            q75 = quantile(x, 0.75)) %>% 
  mutate(whisker_low = q25 - 1.5 * (q75 - q25),
         whisker_high = q75 + 1.5 * (q75 - q25)) %>% 
  pull(whisker_low)

gauss_whisker_high <- xdf %>% 
  summarise(q25 = quantile(x, 0.25),
            q50 = median(x),
            q75 = quantile(x, 0.75)) %>% 
  mutate(whisker_low = q25 - 1.5 * (q75 - q25),
         whisker_high = q75 + 1.5 * (q75 - q25)) %>% 
  pull(whisker_high)

(gauss_whiskers <- c(gauss_whisker_low, gauss_whisker_high))
```

Draw the manually calculated whisker endpoints on the boxplot.  

```{r, viz_fake_box_whiskers}
xdf %>% 
  ggplot(mapping = aes(y = x)) +
  geom_boxplot() +
  geom_hline(yintercept = gauss_whiskers,
             color = 'red', linetype = 'dashed', size = 1.2) +
  scale_x_discrete()
```
 
Why 1.5 times the IQR?  

```{r, check_iqr_multiplier}
xdf %>% 
  mutate(xbins = cut(x,
                     breaks = c(min(x),
                                gauss_whisker_low,
                                gauss_whisker_high,
                                max(x)),
                     include.lowest = TRUE,
                     labels = c("low-outliers", "non-outliers", "high-outliers"))) %>% 
  count(xbins) %>% 
  mutate(percentage_value = 100 * (n / sum(n)))
```

## Conditioning on a categorical

Review the `mpg` data set.  

```{r, mpg_glimpse}
mpg %>% glimpse()
```

Look at the distribution of `hwy` for each level of `drv`.  

```{r, viz_freqpoly_hwy_drv}
mpg %>% 
  ggplot(mapping = aes(x = hwy)) +
  geom_freqpoly(mapping = aes(color = drv,
                              y = stat(density)),
                size = 1.2,
                binwidth = 3) +
  ggthemes::scale_color_calc() +
  theme_bw()
```

Using density plots instead of histograms.  

```{r, viz_hwy_density_drv}
mpg %>% 
  ggplot(mapping = aes(x = hwy)) +
  geom_density(mapping = aes(color = drv),
               size = 1.2) +
  ggthemes::scale_color_calc() +
  theme_bw()
```

Visualize the different **conditional** distributions with a violin plot.  

```{r, viz_hwy_drv_violin}
mpg %>% 
  ggplot(mapping = aes(x = drv, y = hwy)) +
  geom_violin(mapping = aes(fill = drv),
              alpha = 0.35) +
  ggthemes::scale_fill_calc(guide = FALSE) +
  theme_bw()
```


Downside of a violin plot is that we cannot easily see important summary statistics.  

```{r, viz_hwy_drv_violin_box}
mpg %>% 
  ggplot(mapping = aes(x = drv, y = hwy)) +
  geom_violin(mapping = aes(fill = drv),
              alpha = 0.35) +
  geom_boxplot(mapping = aes(color = drv),
               size = 1.15,
               fill = NA) +
  ggthemes::scale_fill_calc(guide = FALSE) +
  ggthemes::scale_color_calc(guide = FALSE) +
  theme_bw()
```

Include the observations via a strip plot to get a sense if the violin density estimate is appropriate.  

```{r, viz_hwy_drv_violin_box_strip}
mpg %>% 
  ggplot(mapping = aes(x = drv, y = hwy)) +
  geom_jitter(alpha = 0.5, size = 3, height = 0, width = 0.2) +
  geom_violin(mapping = aes(fill = drv),
              alpha = 0.35) +
  geom_boxplot(mapping = aes(color = drv),
               size = 1.15,
               fill = NA) +
  ggthemes::scale_fill_calc(guide = FALSE) +
  ggthemes::scale_color_calc(guide = FALSE) +
  theme_bw()
```

### Joy plots (ridge plots)

Please download and install `ggridges` from CRAN.  

```{r, load_ggridges_pkg}
library(ggridges)
```

Flip the `x` and `y` aesthetics with ridge-plots compared to box and violin plots.  

```{r, viz_hwy_drv_joy_1}
mpg %>% 
  ggplot(mapping = aes(x = hwy, y = drv)) +
  geom_density_ridges() +
  theme_bw()
```

We can use the `ggridges` theme.  

```{r, viz_hwy_drv_joy_2}
mpg %>% 
  ggplot(mapping = aes(x = hwy, y = drv)) +
  geom_density_ridges() +
  theme_ridges()
```

Map the `fill` to the `drv` variable.  

```{r, viz_hwy_drv_joy_3}
mpg %>% 
  ggplot(mapping = aes(x = hwy, y = drv)) +
  geom_density_ridges(mapping = aes(fill = drv),
                      alpha = 0.35) +
  ggthemes::scale_fill_calc(guide = FALSE) +
  theme_ridges()
```

Modify the vertical "height" of the densities with the `scale` argument.  

```{r, viz_hwy_drv_joy_4}
mpg %>% 
  ggplot(mapping = aes(x = hwy, y = drv)) +
  geom_density_ridges(mapping = aes(fill = drv),
                      alpha = 0.35,
                      scale = 0.75) +
  ggthemes::scale_fill_calc(guide = FALSE) +
  theme_ridges()
```

Modify the bandwidth or the smoothing of the density estimate.  

```{r, viz_hwy_drv_joy_5}
mpg %>% 
  ggplot(mapping = aes(x = hwy, y = drv)) +
  geom_density_ridges(mapping = aes(fill = drv),
                      alpha = 0.35,
                      bandwidth = 1.) +
  ggthemes::scale_fill_calc(guide = FALSE) +
  theme_ridges()
```

Add quantiles to the joy plot.  

```{r, viz_hwy_drv_joy_6}
mpg %>% 
  ggplot(mapping = aes(x = hwy, y = drv)) +
  geom_density_ridges(mapping = aes(fill = drv),
                      alpha = 0.35,
                      bandwidth = 1.,
                      quantile_lines = TRUE) +
  ggthemes::scale_fill_calc(guide = FALSE) +
  theme_ridges()
```

Separate fill color for the quartile ranges. To make the following we have to use the underlying `stat_` function.  

```{r, viz_hwy_drv_joy_7}
mpg %>% 
  ggplot(mapping = aes(x = hwy, y = drv)) +
  stat_density_ridges(geom = "density_ridges_gradient",
                      mapping = aes(fill = factor(stat(quantile))),
                      calc_ecdf = TRUE,
                      quantiles = 4,
                      quantile_lines = TRUE,
                      bandwidth = 1.) +
  scale_fill_viridis_d("Quartiles") +
  theme_ridges()
```

We can specify our own set of quantiles if we prefer. Such as the middle 80% interval.  

```{r, viz_hwy_drv_joy_8}
mpg %>% 
  ggplot(mapping = aes(x = hwy, y = drv)) +
  geom_density_ridges(mapping = aes(fill = drv),
                      alpha = 0.35,
                      bandwidth = 1.,
                      quantile_lines = TRUE,
                      quantiles = c(0.1, 0.9)) +
  ggthemes::scale_fill_calc(guide = FALSE) +
  theme_ridges()
```

Fill in the 3 intervals with a manual color scale.  

```{r, viz_hwy_drv_joy_9}
mpg %>% 
  ggplot(mapping = aes(x = hwy, y = drv)) +
  stat_density_ridges(geom = "density_ridges_gradient",
                      mapping = aes(fill = factor(stat(quantile))),
                      calc_ecdf = TRUE,
                      quantile_lines = TRUE,
                      quantiles = c(0.1, 0.9),
                      bandwidth = 1.) +
  scale_fill_manual("Probability",
                    values = c("darkred", "grey", "navyblue"),
                    labels = c("(0, 0.1]", "(0.1, 0.9]", "(0.9, 1]")) +
  theme_ridges()
```

To understand if the density is representative of the underlying data, let's include the measruements directly on the graphic.  

```{r, viz_hwy_drv_joy_10}
mpg %>% 
  ggplot(mapping = aes(x = hwy, y = drv)) +
  geom_density_ridges(mapping = aes(fill = drv),
                      alpha = 0.35,
                      bandwidth = 1.,
                      quantile_lines = TRUE,
                      jittered_points = TRUE,
                      rel_min_height = 0.005) +
  ggthemes::scale_fill_calc(guide = FALSE) +
  theme_ridges()
```

We can position the markers below the density plot to create the **raincloud** plot.  

```{r, viz_hwy_drv_joy_11}
mpg %>% 
  ggplot(mapping = aes(x = hwy, y = drv)) +
  geom_density_ridges(mapping = aes(fill = drv),
                      alpha = 0.35,
                      bandwidth = 1.,
                      quantile_lines = TRUE,
                      jittered_points = TRUE,
                      rel_min_height = 0.005,
                      position = 'raincloud') +
  ggthemes::scale_fill_calc(guide = FALSE) +
  theme_ridges()
```

We can modify the points to look more like a rug plot to further help identify high density intervals.  

```{r, viz_hwy_drv_joy_12}
mpg %>% 
  ggplot(mapping = aes(x = hwy, y = drv)) +
  geom_density_ridges(mapping = aes(fill = drv),
                      alpha = 0.35,
                      bandwidth = 1.,
                      quantile_lines = TRUE,
                      jittered_points = TRUE,
                      rel_min_height = 0.005,
                      position = position_points_jitter(width = 0.2, height = 0),
                      point_shape = "|", point_alpha = 0.3, point_size = 3) +
  ggthemes::scale_fill_calc(guide = FALSE) +
  theme_ridges()
```

Joy plots are particularly useful when you are comparing how a distribution "evolves" over time.  

```{r, nebraska_temp_glimpse}
lincoln_weather %>% glimpse()
```

Look how the average or mean temperature per day distribution evolves over the months.  

```{r, nebraska_temp_viz_1}
lincoln_weather %>% 
  ggplot(mapping = aes(x = `Mean Temperature [F]`, y = Month)) +
  geom_density_ridges() +
  theme_ridges()
```

Modifying the relative min height can help with the visual sometimes.  

```{r, nebraska_temp_viz_2}
lincoln_weather %>% 
  ggplot(mapping = aes(x = `Mean Temperature [F]`, y = Month)) +
  geom_density_ridges(rel_min_height = 0.01) +
  theme_ridges()
```

Modify the height to help with the visual.  

```{r, nebraksa_temp_viz_3}
lincoln_weather %>% 
  ggplot(mapping = aes(x = `Mean Temperature [F]`, y = Month)) +
  geom_density_ridges(rel_min_height = 0.01, scale=3) +
  theme_ridges()
```

Fill the ridge area based on the value on the `x` axis.  

```{r, nebraska_temp_viz_4}
lincoln_weather %>% 
  ggplot(mapping = aes(x = `Mean Temperature [F]`, y = Month)) +
  geom_density_ridges_gradient(rel_min_height = 0.01, scale=3,
                               mapping = aes(fill = stat(x))) +
  scale_fill_viridis_c("Temp [F]", option = 'plasma') +
  theme_ridges()
```

Raincloud for the temperatures.  

```{r, nebraska_temp_viz_5}
lincoln_weather %>% 
  ggplot(mapping = aes(x = `Mean Temperature [F]`, y = Month)) +
  geom_density_ridges(rel_min_height = 0.01, scale=3,
                      jittered_points = TRUE,
                      position = 'raincloud',
                      alpha = 0.5) +
  theme_ridges()
```

Please see the documentation [ggridges](https://wilkelab.org/ggridges/index.html) and the [vignette](https://cran.r-project.org/web/packages/ggridges/vignettes/gallery.html)  

## Comparing averages across groups

Look at the distribution of `hwy` per `drv`.  

```{r, viz_hwy_drv_violin_box_strip_2b}
mpg %>% 
  ggplot(mapping = aes(x = drv, y = hwy)) +
  geom_jitter(alpha = 0.5, size = 3, height = 0, width = 0.2) +
  geom_violin(mapping = aes(fill = drv),
              alpha = 0.35) +
  geom_boxplot(mapping = aes(color = drv),
               size = 1.15,
               fill = NA) +
  ggthemes::scale_fill_calc(guide = FALSE) +
  ggthemes::scale_color_calc(guide = FALSE) +
  theme_bw()
```

Calculate the group-level averages.  

```{r, calc_hwy_drv_averages}
mpg %>% 
  group_by(drv) %>% 
  summarise(avg_hwy = mean(hwy))
```

We can override the default or **inherited** data argument from the parent `ggplot()` call. Each geometric layer can have its own data set.  

```{r, viz_avg_hwy_manual}
mpg %>% 
  ggplot(mapping = aes(x = drv, y = hwy)) +
  geom_jitter(alpha = 0.5, size = 3, height = 0, width = 0.2) +
  geom_violin(mapping = aes(fill = drv),
              alpha = 0.35) +
  geom_boxplot(mapping = aes(color = drv),
               size = 1.15,
               fill = NA) +
  # override the data argument
  geom_point(data = mpg %>% 
               group_by(drv) %>% 
               summarise(avg_hwy = mean(hwy)),
             mapping = aes(y = avg_hwy,
                           shape = drv),
             size = 7) +
  ggthemes::scale_fill_calc(guide = FALSE) +
  ggthemes::scale_color_calc(guide = FALSE) +
  scale_shape_discrete(guide = FALSE) +
  theme_bw()
```

We can work with the original data and instruct ggplot2 to summarize via `stat_summary()`.  

```{r, viz_avg_hwy_drv_1}
mpg %>% 
  ggplot(mapping = aes(x = drv, y = hwy)) +
  geom_jitter(alpha = 0.5, size = 3, height = 0, width = 0.2) +
  geom_violin(mapping = aes(fill = drv),
              alpha = 0.35) +
  geom_boxplot(mapping = aes(color = drv),
               size = 1.15,
               fill = NA) +
  # summarize the data via stat_summary()
  stat_summary(geom = 'point',
               fun = 'mean',
               mapping = aes(shape = drv),
               size = 7) +
  ggthemes::scale_fill_calc(guide = FALSE) +
  ggthemes::scale_color_calc(guide = FALSE) +
  scale_shape_discrete(guide = FALSE) +
  theme_bw()
```

When showing the group-level averages sometimes it's useful to not show the individual observations.  

```{r, viz_avg_hwy_drv_2}
mpg %>% 
  ggplot(mapping = aes(x = drv, y = hwy)) +
  # geom_jitter(alpha = 0.5, size = 3, height = 0, width = 0.2) +
  geom_violin(mapping = aes(fill = drv),
              alpha = 0.35) +
  geom_boxplot(mapping = aes(color = drv),
               size = 1.15,
               fill = NA) +
  # summarize the data via stat_summary()
  stat_summary(geom = 'point',
               fun = 'mean',
               mapping = aes(shape = drv),
               size = 7) +
  ggthemes::scale_fill_calc(guide = FALSE) +
  ggthemes::scale_color_calc(guide = FALSE) +
  scale_shape_discrete(guide = FALSE) +
  theme_bw()
```

But how confident are we in the average?  

What is the **standard error** on the sample average?  

$$ 
se = \sqrt{ \frac{Var(x)}{N} }
$$

Calculate the standard error on the average and calculate the 95% confidence interval on the average.  

```{r, calc_hyw_avg_se}
mpg %>% 
  group_by(drv) %>% 
  summarise(num_rows = n(),
            avg_hwy = mean(hwy),
            sd_hwy = sd(hwy)) %>% 
  ungroup() %>% 
  mutate(avg_hwy_se = sd_hwy / sqrt(num_rows)) %>% 
  mutate(avg_hwy_lwr = avg_hwy - 2 * avg_hwy_se,
         avg_hwy_upr = avg_hwy + 2 * avg_hwy_se) %>% 
  select(drv, avg_hwy_lwr, avg_hwy, avg_hwy_upr)
```

We can visualize the sample average and the confidence interval on the sample average with the `geom_pointrange()`.  

```{r, calc_hyw_avg_se_summary}
drv_group_summary <- mpg %>% 
  group_by(drv) %>% 
  summarise(num_rows = n(),
            avg_hwy = mean(hwy),
            sd_hwy = sd(hwy)) %>% 
  ungroup() %>% 
  mutate(avg_hwy_se = sd_hwy / sqrt(num_rows)) %>% 
  mutate(avg_hwy_lwr = avg_hwy - 2 * avg_hwy_se,
         avg_hwy_upr = avg_hwy + 2 * avg_hwy_se)
```

Let's override the default data argument and manually show the `geom_pointrange()`.  

```{r, viz_hwy_avg_comparison_1}
mpg %>% 
  ggplot(mapping = aes(x = drv, y = hwy)) +
  # geom_jitter(alpha = 0.5, size = 3, height = 0, width = 0.2) +
  geom_violin(mapping = aes(fill = drv),
              alpha = 0.35) +
  geom_boxplot(mapping = aes(color = drv),
               size = 1.15,
               fill = NA) +
  # manually override the data argument
  geom_pointrange(data = drv_group_summary,
                  mapping = aes(y = avg_hwy,
                                ymin = avg_hwy_lwr,
                                ymax = avg_hwy_upr,
                                shape = drv),
                  color = 'black') +
  ggthemes::scale_fill_calc(guide = FALSE) +
  ggthemes::scale_color_calc(guide = FALSE) +
  scale_shape_discrete(guide = FALSE) +
  theme_bw()
```

We can use `stat_summary()` to not only show the sample average, but we can show the confidence interval on the sample average. We must specify a function to apply to the entire data set within `stat_summary()`. By default `stat_summary()` with `fun.data = 'mean_se'` shows the $\pm$1 standard error interval (68% confidence interval) so we must override this.  

```{r, viz_hwy_drv_group_avg_se_2}
mpg %>% 
  ggplot(mapping = aes(x = drv, y = hwy)) +
  # geom_jitter(alpha = 0.5, size = 3, height = 0, width = 0.2) +
  geom_violin(mapping = aes(fill = drv),
              alpha = 0.35) +
  geom_boxplot(mapping = aes(color = drv),
               size = 1.15,
               fill = NA) +
  stat_summary(fun.data = 'mean_se',
               mapping = aes(shape = drv),
               fun.args = list(mult = 2)) +
  ggthemes::scale_fill_calc(guide = FALSE) +
  ggthemes::scale_color_calc(guide = FALSE) +
  scale_shape_discrete(guide = FALSE) +
  theme_bw()
```

We can use `stat_summary()` to summarize a data set however we wish.  

```{r, make_a_quantile_function}
my_quantile_0.1 <- function(x, p=0.1){quantile(x, p)}

my_quantile_0.25 <- function(x, p=0.25){quantile(x, p)}

my_quantile_0.75 <- function(x, p=0.75){quantile(x, p)}

my_quantile_0.9 <- function(x, p=0.9){quantile(x, p)}
```

Instead of showing a boxplot, use `stat_summary()` to display the middle 50% interval (the box) and display the middle 80% interval.  

```{r, viz_hwy_drv_avg_summary_1}
mpg %>% 
  ggplot(mapping = aes(x = drv, y = hwy)) +
  geom_violin(mapping = aes(fill = drv), alpha = 0.25) +
  stat_summary(geom = 'linerange',
               fun.min = 'min', fun.max = 'max',
               mapping = aes(group = drv),
               color = 'grey30', size = 0.75) +
  stat_summary(geom = 'linerange',
               fun.min = my_quantile_0.1,
               fun.max = my_quantile_0.9,
               mapping = aes(group = drv,
                             color = drv),
               size = 2.5) +
  stat_summary(geom = 'linerange',
               fun.min = my_quantile_0.25,
               fun.max = my_quantile_0.75,
               mapping = aes(group = drv,
                             color = drv),
               size = 7.5) +
  stat_summary(fun.data = "mean_se",
               color = 'black',
               fun.args = list(mult = 2)) +
  ggthemes::scale_fill_calc(guide = FALSE) +
  ggthemes::scale_color_calc(guide = FALSE) +
  theme_bw()
```

When I make these kind of figures I use an inline function.  

```{r, viz_hwy_drv_avg_summary_2}
mpg %>% 
  ggplot(mapping = aes(x = drv, y = hwy)) +
  geom_violin(mapping = aes(fill = drv), alpha = 0.25) +
  stat_summary(geom = 'linerange',
               fun.min = 'min', fun.max = 'max',
               mapping = aes(group = drv),
               color = 'grey30', size = 0.75) +
  stat_summary(geom = 'linerange',
               fun.min = function(x){quantile(x, 0.1)},
               fun.max = function(x){quantile(x, 0.9)},
               mapping = aes(group = drv,
                             color = drv),
               size = 2.5) +
  stat_summary(geom = 'linerange',
               fun.min = function(x){quantile(x, 0.25)},
               fun.max = function(x){quantile(x, 0.75)},
               mapping = aes(group = drv,
                             color = drv),
               size = 7.5) +
  stat_summary(fun.data = "mean_se",
               color = 'black',
               fun.args = list(mult = 2)) +
  ggthemes::scale_fill_calc(guide = FALSE) +
  ggthemes::scale_color_calc(guide = FALSE) +
  theme_bw()
```

If we are only interested in the variation in the group-level averages:  

```{r, viz_hwy_drv_group_avg_3}
mpg %>% 
  ggplot(mapping = aes(x = drv, y = hwy)) +
  stat_summary(fun.data = 'mean_se',
               mapping = aes(color = drv,
                             shape = drv),
               fun.args = list(mult = 2),
               size = 1.75) +
  ggthemes::scale_color_calc(guide = FALSE) +
  scale_shape_discrete(guide = FALSE) +
  theme_bw()
```

### T-test for the Nebraska temperature data

```{r, nebraska_avg_temp_t_test_viz_1}
lincoln_weather %>% 
  ggplot(mapping = aes(x = Month, y = `Mean Temperature [F]`)) +
  stat_summary(fun.data = 'mean_se',
               fun.args = list(mult = 2)) +
  theme_bw()
```

Check the order of the levels.  

```{r, check_nebraska_levels_months}
lincoln_weather %>% 
  pull(Month) %>% 
  levels()
```

We can reverse the level order with `forcats`.  

```{r, nebraska_avg_temp_t_test_viz_2}
lincoln_weather %>% 
  mutate(Month = forcats::fct_rev(Month)) %>% 
  ggplot(mapping = aes(x = Month, y = `Mean Temperature [F]`)) +
  stat_summary(fun.data = 'mean_se',
               fun.args = list(mult = 2)) +
  theme_bw()
```

Easier to **flip** the coordinate system to put the months on the y-axis.  

```{r, nebraska_avg_temp_t_test_viz_3}
lincoln_weather %>% 
  # mutate(Month = forcats::fct_rev(Month)) %>% 
  ggplot(mapping = aes(x = Month, y = `Mean Temperature [F]`)) +
  stat_summary(fun.data = 'mean_se',
               fun.args = list(mult = 2)) +
  coord_flip() +
  theme_bw()
```

### Visual t-test for the iris data set

```{r, iris_glimpse}
iris %>% glimpse()
```

Calculate the average `Sepal.Length` per `Species` and check if there are differences in the average across `Species`.  

```{r, iris_viz_sepal_length_t_test_1}
iris %>% 
  ggplot(mapping = aes(x = Species, y = Sepal.Length)) +
  stat_summary(fun.data = 'mean_se',
               mapping = aes(color = Species),
               fun.args = list(mult = 2),
               size = 2) +
  ggthemes::scale_color_colorblind(guide = FALSE) +
  theme_bw()
```

## Continuous variable relationships

Looked at the correlation coefficient between the `Sepal` variables.  

```{r, iris_sepal_corr}
iris %>% 
  select(starts_with("Sepal.")) %>% 
  cor()
```

Looked at the correlation coefficient **grouped by** `Species`.  

```{r, iris_sepal_corr_species}
iris %>% 
  select(Species, starts_with("Sepal.")) %>% 
  group_by(Species) %>% 
  group_modify( ~ as.data.frame(cor(.x)))
```


Visualizing correlation plots can streamline understanding the linear relationship across many **pairs** of variables.  

```{r, iris_cor_mat}
iris %>% 
  purrr::keep(is.numeric) %>% 
  cor()
```

### Sonar data

Need the `mlbench` package.  

```{r, read_sonar_data}
data("Sonar", package = 'mlbench')
```

```{r, sonar_dim_check}
Sonar %>% dim()
```

```{r, check_sonar_datatypes}
Sonar %>% purrr::map_chr(class)
```

Check the correlation between every pair of variables.  

```{r, sonar_cor_mat_check}
Sonar %>% 
  purrr::keep(is.numeric) %>% 
  cor() %>% 
  dim()
```

Visualize the correlation plot using `corrplot`.  

```{r, load_corrplot_pkg}
library(corrplot)
```

`corrplot` has a main function `corrplot()` that requires a correlation matrix as input.  

```{r, sonar_corrplot_viz_1}
Sonar %>% 
  purrr::keep(is.numeric) %>% 
  cor() %>% 
  corrplot::corrplot()
```

When there are many variables I prefer to use the squares instead of circles.  

```{r, sonar_corrplot_viz_2}
Sonar %>% 
  purrr::keep(is.numeric) %>% 
  cor() %>% 
  corrplot::corrplot(method = 'square')
```

Can just show a upper triangular plot.  

```{r, sonar_corrplot_viz_3}
Sonar %>% 
  purrr::keep(is.numeric) %>% 
  cor() %>% 
  corrplot::corrplot(method = 'square', type = 'upper')
```

We can reorganize the variables based on **groups** or **clusters** of variables.  

```{r, sonar_corrplot_viz_4}
Sonar %>% 
  purrr::keep(is.numeric) %>% 
  cor() %>% 
  corrplot::corrplot(method = 'square', order = 'hclust')
```

By default the complete linkage is used, but I prefer the Ward method.  

```{r, sonar_corrplot_viz_5}
Sonar %>% 
  purrr::keep(is.numeric) %>% 
  cor() %>% 
  corrplot::corrplot(method = 'square', order = 'hclust', hclust.method = 'ward.D2')
```

Lastly we can draw rectangles around the clusters.  

```{r, sonar_corrplot_viz_6}
Sonar %>% 
  purrr::keep(is.numeric) %>% 
  cor() %>% 
  corrplot::corrplot(method = 'square', order = 'hclust', hclust.method = 'ward.D2',
                     addrect=8)
```

With smaller data sets like `iris` we can show the correlation coefficient value directly.  

```{r, iris_viz_corrplot}
iris %>% 
  purrr::keep(is.numeric) %>% 
  cor() %>% 
  corrplot::corrplot.mixed()
```

