---
title: "CMPINF 2130 Summer 2021 - Week 08"
subtitle: "pairsplots, 2D histograms"
author: "Dr. Joseph P. Yurko"
date: "July 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages

```{r, load_packages}
library(tidyverse)

library(GGally)
```

## Iris data

`iris` has 4 continuous variables.  

```{r, check_iris_types}
iris %>% purrr::map_chr(class)
```

Correlation plot showing the correlation coefficient for all numeric variables.  

```{r, corrplot_iris_viz}
iris %>% 
  purrr::keep(is.numeric) %>% 
  cor() %>% 
  corrplot::corrplot(method = 'number')
```

Correlation coefficient only tells us linear relationships. It's always important to visualize the data directly. Do not only rely on the correlatioin coefficient to understand the relationship between variables.  

A scatter plot can help us understand the relationship or **joint** behavior of two continuous variables.  

```{r, iris_sepal_scatter_1}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point() +
  theme_bw()
```

A **pairsplot** creates all pairs of scatter plots for us. Sometimes this is called a (pairs) plot matrix.  

The base R `pairs()` function creates the pairsplot.  

```{r, viz_iris_base_pairs}
iris %>% 
  purrr::keep(is.numeric) %>% 
  pairs()
```

`pairs()` also works with categorical variables.  

```{r, viz_iris_base_pairs_2}
iris %>% 
  pairs()
```

`GGally` has a ggplot2 style version of base `R` `pairs()` named `ggpairs()`.  

Let's first look at the `ggpairs()` of the continuous variables only.  

```{r, ggpairs_iris_1}
iris %>% 
  purrr::keep(is.numeric) %>% 
  ggpairs()
```


Let's hide the progress by setting `progress = FALSE` in the `ggpairs()` call.  

```{r, ggpairs_iris_2}
iris %>% 
  purrr::keep(is.numeric) %>% 
  ggpairs(progress = FALSE)
```

`ggpairs()` produces a ggplot object, so we can set the theme just like we did with "regular" ggplot objects.  

```{r, ggpairs_iris_3}
iris %>% 
  purrr::keep(is.numeric) %>% 
  ggpairs(progress = FALSE) +
  theme_bw()
```

We can use a histogram for the main diagonal instead of the kernel destiny estimate.  

```{r, ggpairs_iris_4}
iris %>% 
  purrr::keep(is.numeric) %>% 
  ggpairs(progress = FALSE,
          diag = list(continuous = 'barDiag')) +
  theme_bw()
```


We can specify the number of bins for the marginal histograms.  

```{r, ggpairs_iris_5}
iris %>% 
  purrr::keep(is.numeric) %>% 
  ggpairs(progress = FALSE,
          diag = list(continuous = wrap('barDiag', bins=25))) +
  theme_bw()
```

We can use the `columns` argument in `ggpairs()` to specify the columns/variables to display.  

```{r, ggpairs_iris_6}
iris %>% 
  ggpairs(columns = 1:4,
          progress = FALSE,
          diag = list(continuous = wrap('barDiag', bins=25))) +
  theme_bw()
```

We must know the order of the variables to use the `columns` argument.  

```{r, ggpairs_iris_7}
iris %>% names()
```

We can always look at the scatter plot directly to check.  

```{r, check_a_scatter_plot}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Petal.Width)) +
  geom_point() +
  theme_bw()
```


`ggpairs()` also works with categorical variables.  

```{r, ggpairs_iris_8}
iris %>% 
  ggpairs(progress = FALSE,
          diag = list(continuous = wrap('barDiag', bins=25))) +
  theme_bw()
```


We can use stripplots for the lower diagonal for the "combination" between continuous and categorical to be consistent with the continuous scatter plots.  

```{r, ggpairs_iris_9}
iris %>% 
  ggpairs(progress = FALSE,
          diag = list(continuous = wrap('barDiag', bins=25)),
          lower = list(combo = 'dot_no_facet')) +
  theme_bw()
```

Usually when I work with pairs plots I color by a categorical rather than include it in the subplots.  

```{r, ggpairs_iris_10}
iris %>% 
  ggpairs(columns = 1:4,
          progress = FALSE,
          diag = list(continuous = wrap('barDiag', bins=25)),
          mapping = aes(color = Species)) +
  theme_bw()
```

If we do not specify the `diag` we will get the default density plots along the main diagonal.  

```{r, ggpairs_iris_11}
iris %>% 
  ggpairs(columns = 1:4,
          progress = FALSE,
          mapping = aes(color = Species)) +
  theme_bw()
```


## Joint distributions

What if there are many observations right on top of each other or very close together? Essentially, how do we handle a "busy" scatter plot?  


For example, let's look at the scatter plot between the Sepal variables.  

```{r, sepal_scatter_1}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(size = 3.5) +
  theme_bw()
```

Transparency can help identify "high density" regions.  

```{r, sepal_scatter_2}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(size = 3.5, alpha=0.25) +
  theme_bw()
```

Or we can count the number of observations per combination.  

```{r, sepal_scatter_3}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_count() +
  theme_bw()
```

But what if we have more data? Or what there are many tightly packed observations, but a large number of unique values?  

For example, let's work with the `diamonds` data set.  

```{r, glimspse_diamonds_data}
diamonds %>% dplyr::glimpse()
```

Look at the scatter plot between `price` and `carat`.  

```{r, viz_carat_price_scatter}
diamonds %>% 
  ggplot(mapping = aes(x = carat, y = price)) +
  geom_point() +
  theme_bw()
```

Raise transparency (reduce `alpha`).  

```{r, viz_carat_price_scatter_2}
diamonds %>% 
  ggplot(mapping = aes(x = carat, y = price)) +
  geom_point(alpha = 0.005) +
  theme_bw()
```


If we count...we can't really see anything...  

```{r, viz_cara_price_scatter_count}
diamonds %>% 
  ggplot(mapping = aes(x = carat, y = price)) +
  geom_count() +
  theme_bw()
```

We want to know where the data are dense vs less concentrated. We can use 2D histogram to represent the **joint distribution** between two continuous variables.  

```{r, viz_bin2d_carat_price}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_bin_2d() +
  theme_bw()
```

Essentially `geom_bin_2d()` performs two `cut()` functions behind the scenes and then `count()` by the combinations of the discretized continuous variables.  

We can raise or lower the number of bins.  

```{r, viz_bin2d_carat_price_2}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_bin_2d(bins = 11) +
  theme_bw()
```

Because the bin is defined based on 2 variables we can use two separate bin sizes.  

```{r, viz_bin2d_carat_price_3}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_bin_2d(bins = c(101, 5)) +
  theme_bw()
```

I tend to start with 11 bins in each direction.  

```{r, viz_bin2d_carat_price_4}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_bin_2d(bins = 11) +
  scale_fill_viridis_c() +
  theme_bw()
```

Let's now make a 2D histogram for the `price` and `carat` joint distribution.  

```{r, viz_bin2d_carat_price_5}
diamonds %>% 
  ggplot(mapping = aes(x = carat, y = price)) +
  geom_bin_2d(bins = 11) +
  scale_fill_viridis_c() +
  theme_bw()
```


Try the default number of bins.  

```{r, viz_bin2d_carat_price_6}
diamonds %>% 
  ggplot(mapping = aes(x = carat, y = price)) +
  geom_bin_2d() +
  scale_fill_viridis_c() +
  theme_bw()
```

Alternatively instead of the rectangles we can use hexagons to represent the 2D histogram.  

```{r, viz_hex_carat_price_1}
diamonds %>% 
  ggplot(mapping = aes(x = carat, y = price)) +
  geom_hex() +
  scale_fill_viridis_c() +
  theme_bw()
```


### Conditioning

We can condition a scatter plot on a categorical variable. First look at iris sepal variables per Species.  

```{r, iris_scatter_per_species}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point() +
  facet_wrap(~Species) +
  theme_bw()
```

We can visualize the joint distribution as a 2D histogram for separate levels of a categorical.  

```{r, iris_hex_per_species}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_hex(bins = 11) +
  facet_wrap(~Species) +
  scale_fill_viridis_c() +
  theme_bw()
```

Use the `cut` variable as the grouping categorical variable in `diamonds`.  

```{r, check_cut_levels}
diamonds %>% dplyr::count(cut)
```


Condition the hexbin plot of `price` and `carat` on `cut`.  

```{r, viz_carat_price_hex_cut}
diamonds %>% 
  ggplot(mapping = aes(x = carat, y = price)) +
  geom_hex() +
  facet_wrap(~cut) +
  scale_fill_viridis_c() +
  theme_bw()
```

We can scale the fill color relative to the max count per faceting variable group.  

```{r, viz_carat_price_hex_cut_2}
diamonds %>% 
  ggplot(mapping = aes(x = carat, y = price)) +
  geom_hex(mapping = aes(fill = stat(ncount))) +
  facet_wrap(~cut) +
  scale_fill_viridis_c() +
  theme_bw()
```

Check by filtering to the `cut == 'Fair'`.  

```{r, viz_carat_price_hex_cut_3}
diamonds %>% 
  dplyr::filter(cut == "Fair") %>% 
  ggplot(mapping = aes(x = carat, y = price)) +
  geom_hex() +
  facet_wrap(~cut) +
  scale_fill_viridis_c() +
  theme_bw()
```

We can also us a kernel density estimate to smooth out the counts of a histogram.  

Let's first see the 2D density estimate for iris.  

```{r, iris_sepal_2d_density}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_density_2d() +
  theme_bw()
```

We can overlay the scatter plot to confirm the contour behavior.  

```{r, iris_sepal_2d_density_2}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(size = 3, alpha = 0.25) +
  geom_density_2d() +
  theme_bw()
```

Alternatively we can compare the 2D histogram with the 2D density estimate.  

```{r, iris_sepal_2d_density_3}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_hex(bins=11) +
  geom_density_2d(color = 'darkorange', size = 1.25) +
  scale_fill_viridis_c() +
  theme_bw()
```


We can also use a filled in contour plot if we would like.  

```{r, iris_sepal_2d_density_4}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_density_2d_filled() +
  theme_bw()
```

We can also condition the 2D density estimate on a categorical variable.  

```{r, iris_sepal_2d_density_5}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_density_2d_filled() +
  facet_wrap(~Species) +
  theme_bw()
```

And so we may instead prefer to use a relative fill scale to focus on the combinations with the highest relative density in each facet.  

```{r, iris_sepal_2d_density_6}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_density_2d_filled(contour_var = 'ndensity') +
  facet_wrap(~Species) +
  theme_bw()
```

Alternatively, we can overlay densities on a single plot colored by the categorical variable.  

```{r, iris_sepal_2d_density_7}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_density_2d(mapping = aes(color = Species),
                  size = 1.1) +
  ggthemes::scale_color_colorblind() +
  theme_bw()
```

Overlaying density contours by coloring with categoricals is popular in pairs plots.  

```{r, iris_pairs_plots_again}
iris %>% 
  ggpairs(columns = 1:4,
          progress = FALSE,
          mapping = aes(color = Species)) +
  theme_bw()
```

Instead use a 2D density estimate, but it is important to use the `ndensity` to compare the relative heights across the categorical groups.  

```{r, iris_pairs_plot_ndensity_2d}
iris %>% 
  ggpairs(columns = 1:4,
          progress = FALSE,
          mapping = aes(color = Species),
          upper = list(continuous = wrap('density', 
                                         contour_var = 'ndensity',
                                         adjust = 1.5))) +
  theme_bw()
```

There are other options we can consider for how to display the diagonal and off-diagonal subplots within a pairs plot. Please see the `ggpairs()` documentation to learn more.  

[ggpairs main reference](https://ggobi.github.io/ggally/reference/ggpairs.html)  

[ggpairs function types](https://ggobi.github.io/ggally/articles/ggally_plots.html)  
