---
title: "CMPINF 2130 Summer 2021 - Week 08"
subtitle: "correlation plots per group"
author: "Dr. Joseph P. Yurko"
date: "July 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages

```{r, load_packages}
library(tidyverse)

library(corrr)
```

`corrr` is a package dedicated to working with, manipulating, and visualizing correlation matrices.  

## Iris data

```{r, calc_iris_corrr_matrix}
iris_corrr <- iris %>% 
  purrr::keep(is.numeric) %>% 
  corrr::correlate()
```

Contrast this with the behavior from `cor()` function from base `R`.  

```{r, calc_iris_cor_base}
iris_cor_base <- iris %>% 
  purrr::keep(is.numeric) %>% 
  cor()
```


Display the correlation matrix from base R.  

```{r, show_cor_mat_iris}
iris_cor_base
```

Display the `correlate()` correlation matrix.  

```{r, show_corrr_mat_iris}
iris_corrr
```

Change the diagonal behavior.  

```{r, check_diag_corrr_behavior}
iris %>% 
  purrr::keep(is.numeric) %>% 
  correlate(diagonal = 1)
```


Visualize the correlation dataframe using `rplot()`.  

```{r, viz_corrr_rplot}
iris_corrr %>% rplot()
```

`corrr` is a network style graph for visualizing the correlations.  

```{r, viz_corrr_network}
iris_corrr %>% network_plot()
```

`corrr` also allows reshaping the correlation dataframe into a long-format.  

```{r, reshape_stretch_corrr}
iris_corrr %>% stretch()
```

The long-format is especially useful for working with grouped correlation dataframes. Let's create the grouped correlation dataframe for each level of `Species`.  

```{r, make_grouped_corrr_1}
iris %>% 
  group_by(Species) %>% 
  tidyr::nest() %>% 
  mutate(cor_wf = purrr::map(data, correlate)) %>% 
  mutate(cor_lf = purrr::map(cor_wf, stretch)) %>% 
  select(Species, cor_lf) %>% 
  tidyr::unnest(cor_lf)
```

We can now visualize the correlation coefficient with `ggplot2` using facets to display the levels of `Species`.  

```{r, viz_corr_group_1}
iris %>% 
  group_by(Species) %>% 
  tidyr::nest() %>% 
  mutate(cor_wf = purrr::map(data, correlate)) %>% 
  mutate(cor_lf = purrr::map(cor_wf, stretch)) %>% 
  select(Species, cor_lf) %>% 
  tidyr::unnest(cor_lf) %>% 
  ungroup() %>% 
  ggplot(mapping = aes(x = x, y = y)) +
  geom_tile(mapping = aes(fill = r),
            color = 'white') +
  facet_wrap(~Species) +
  labs(x = '', y ='') +
  theme_bw()
```

Modify the correlation plot fill and legend.  

```{r, viz_corr_group_2}
iris %>% 
  group_by(Species) %>% 
  tidyr::nest() %>% 
  mutate(cor_wf = purrr::map(data, correlate)) %>% 
  mutate(cor_lf = purrr::map(cor_wf, stretch)) %>% 
  select(Species, cor_lf) %>% 
  tidyr::unnest(cor_lf) %>% 
  ungroup() %>% 
  ggplot(mapping = aes(x = x, y = y)) +
  geom_tile(mapping = aes(fill = r),
            color = 'white') +
  facet_wrap(~Species) +
  scale_fill_gradient2('corr',
                       low = 'indianred2',
                       mid = 'white',
                       high = 'skyblue1',
                       midpoint = 0,
                       limits = c(-1, 1)) +
  labs(x = '', y ='') +
  theme_bw()
```

Rotate the x-axis label text.  

```{r, viz_corr_group_3}
iris %>% 
  group_by(Species) %>% 
  tidyr::nest() %>% 
  mutate(cor_wf = purrr::map(data, correlate)) %>% 
  mutate(cor_lf = purrr::map(cor_wf, stretch)) %>% 
  select(Species, cor_lf) %>% 
  tidyr::unnest(cor_lf) %>% 
  ungroup() %>% 
  ggplot(mapping = aes(x = x, y = y)) +
  geom_tile(mapping = aes(fill = r),
            color = 'white') +
  facet_wrap(~Species) +
  scale_fill_gradient2('corr',
                       low = 'indianred2',
                       mid = 'white',
                       high = 'skyblue1',
                       midpoint = 0,
                       limits = c(-1, 1)) +
  labs(x = '', y ='') +
  theme_bw() +
  theme(axis.text.x = element_text(angle=90, vjust=0.5))
```

Turn off or make the correlation function work "quietly".  

```{r, viz_corr_group_4}
iris %>% 
  group_by(Species) %>% 
  tidyr::nest() %>% 
  mutate(cor_wf = purrr::map(data, correlate, quiet=TRUE)) %>% 
  mutate(cor_lf = purrr::map(cor_wf, stretch)) %>% 
  select(Species, cor_lf) %>% 
  tidyr::unnest(cor_lf) %>% 
  ungroup() %>% 
  ggplot(mapping = aes(x = x, y = y)) +
  geom_tile(mapping = aes(fill = r),
            color = 'white') +
  facet_wrap(~Species) +
  scale_fill_gradient2('corr',
                       low = 'indianred2',
                       mid = 'white',
                       high = 'skyblue1',
                       midpoint = 0,
                       limits = c(-1, 1)) +
  labs(x = '', y ='') +
  theme_bw() +
  theme(axis.text.x = element_text(angle=90, vjust=0.5))
```

I like to dispaly the correlation coefficient as text or to annotate the plot.  

```{r, viz_corr_group_5}
iris %>% 
  group_by(Species) %>% 
  tidyr::nest() %>% 
  mutate(cor_wf = purrr::map(data, correlate, quiet=TRUE)) %>% 
  mutate(cor_lf = purrr::map(cor_wf, stretch)) %>% 
  select(Species, cor_lf) %>% 
  tidyr::unnest(cor_lf) %>% 
  ungroup() %>% 
  ggplot(mapping = aes(x = x, y = y)) +
  geom_tile(mapping = aes(fill = r),
            color = 'white') +
  geom_text(mapping = aes(label = r)) +
  facet_wrap(~Species) +
  scale_fill_gradient2('corr',
                       low = 'indianred2',
                       mid = 'white',
                       high = 'skyblue1',
                       midpoint = 0,
                       limits = c(-1, 1)) +
  labs(x = '', y ='') +
  theme_bw() +
  theme(axis.text.x = element_text(angle=90, vjust=0.5))
```

The `corrr` package has a helper function called `fashion()` to make the text display "pretty". I prefer to use the base R `round()` function to round the value.  

```{r, viz_corrr_group_6}
iris %>% 
  group_by(Species) %>% 
  tidyr::nest() %>% 
  mutate(cor_wf = purrr::map(data, correlate, quiet=TRUE, diagonal=1)) %>% 
  mutate(cor_lf = purrr::map(cor_wf, stretch)) %>% 
  select(Species, cor_lf) %>% 
  tidyr::unnest(cor_lf) %>% 
  ungroup() %>% 
  ggplot(mapping = aes(x = x, y = y)) +
  geom_tile(mapping = aes(fill = r),
            color = 'white') +
  geom_text(mapping = aes(label = round(r, 2))) +
  facet_wrap(~Species) +
  scale_fill_gradient2('corr',
                       low = 'indianred2',
                       mid = 'white',
                       high = 'skyblue1',
                       midpoint = 0,
                       limits = c(-1, 1)) +
  labs(x = '', y ='') +
  theme_bw() +
  theme(axis.text.x = element_text(angle=90, vjust=0.5))
```

## Sonar data

Caclculate the correlation matrix for the continuous variables in `Sonar` and visualize as a network style graphic.  

```{r, read_in_sonar}
data("Sonar", package = 'mlbench')
```

Check the dimensionality.  

```{r, check_sonar_dims}
Sonar %>% dim()
```

Check the data types.  

```{r, check_sonar_class}
Sonar %>% purrr::map_chr(class)
```

Visualize the network style correlation plot.  

```{r, viz_sonar_corr_network}
Sonar %>% 
  purrr::keep(is.numeric) %>% 
  correlate(quiet=TRUE) %>% 
  network_plot(legend = FALSE)
```

Lastly, let's calculate the correlation matrix per group of `Class`.  

```{r, sonar_corr_group}
Sonar %>% 
  group_by(Class) %>% 
  tidyr::nest() %>% 
  mutate(cor_wf = purrr::map(data, correlate, quiet=TRUE, diagonal=1)) %>% 
  mutate(cor_lf = purrr::map(cor_wf, stretch)) %>% 
  select(Class, cor_lf) %>% 
  tidyr::unnest(cor_lf) %>% 
  ungroup() %>% 
  ggplot(mapping = aes(x = x, y = y)) +
  geom_tile(mapping = aes(fill = r), color='white') +
  facet_wrap(~Class, labeller = "label_both") +
  scale_fill_gradient2('corr',
                       low = 'indianred2',
                       mid='white',
                       high='skyblue1',
                       midpoint=0,
                       limits = c(-1,1)) +
  labs(x='',y='') +
  theme_bw() +
  theme(axis.text.x = element_text(angle=90, vjust=0.5))
```

Use string manipulation from the `stringr` package to extract the number for the variable.  

```{r, sonar_corr_group_2}
Sonar %>% 
  group_by(Class) %>% 
  tidyr::nest() %>% 
  mutate(cor_wf = purrr::map(data, correlate, quiet=TRUE, diagonal=1)) %>% 
  mutate(cor_lf = purrr::map(cor_wf, stretch)) %>% 
  select(Class, cor_lf) %>% 
  tidyr::unnest(cor_lf) %>% 
  ungroup() %>% 
  mutate(x_number = stringr::str_extract(x, "\\d+"),
         y_number = stringr::str_extract(y, '\\d+')) %>% 
  mutate(x_number = as.numeric(x_number),
         y_number = as.numeric(y_number)) %>% 
  ggplot(mapping = aes(x = as.factor(x_number), y = as.factor(y_number))) +
  geom_tile(mapping = aes(fill = r), color='white') +
  facet_wrap(~Class, labeller = "label_both") +
  scale_fill_gradient2('corr',
                       low = 'indianred2',
                       mid='white',
                       high='skyblue1',
                       midpoint=0,
                       limits = c(-1,1)) +
  labs(x='',y='') +
  theme_bw() +
  theme(axis.text.x = element_text(angle=90, vjust=0.5))
```

