---
title: "CMPINF 2130 Summer 2021 - Week 08"
subtitle: 'Composite graphics'
author: "Dr. Joseph P. Yurko"
date: "July 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages

```{r, load_packages}
library(tidyverse)

library(ggExtra)
```

## Iris data

Create a scatter plot between the Sepal variables.  

```{r, iris_sepal_scatter}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(alpha = 0.5, size = 3) +
  theme_bw()
```

Creating a composite graphic such as a scatter plot combined with histograms is challenging. We will use the `ggMarginal()` function from `ggExtra` to create the **composite graphic** or the *marginal+scatter* plot.  

First, let's assign the scatter plot to a variable.  

```{r, make_scatter_plot_object}
p <- iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(alpha = 0.5, size = 3) +
  theme_bw()
```


We can visualize the plot by printing or just displaying it.  

```{r, view_plot_object_1}
print( p )
```

Or:  

```{r, view_plot_object_2}
p
```

Modify the scatter plot by combining or adding the marginal histograms outside of the graphical window.  

```{r, add_in_hist_1}
p1 <- ggMarginal(p, type = 'histogram')
```

Print `p1` to visualize the marginal + scatter plot.  

```{r, viz_marginal_p1}
print( p1 )
```

We could instead use kernel density estimates for the marginal distributions.  

```{r, add_in_dens_1}
p2 <- ggMarginal(p, type = 'density')
```

Visualize the marginal + scatter plot again but with kde.  

```{r, viz_marginal_p2}
print( p2 )
```

If we are interested in the marginal summary statistics we can use boxplots outside the graphical window.  

```{r, add_in_box_1}
p3 <- ggMarginal(p, type = 'boxplot')
```


Visualize again.  

```{r, viz_marginal_p3}
print( p3 )
```

We could instead use a 2D joint density estimate to represent the joint distribution.  

```{r, make_2d_density_plot}
sepal_2d <- iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_density_2d(size = 1.1) +
  geom_point(alpha = 0.5, size = 3) +
  theme_bw()

print( sepal_2d )
```

Add in the marginal distributions via kde plots.  

```{r, add_marg_2d_sepal}
sepal_2d_marg <- ggMarginal(sepal_2d, type = 'density')
```

Visualize.  

```{r, viz_marg_2d_sepal}
print( sepal_2d_marg )
```

We can also color by a categorical with the joint distribution **and** the marginal.  

```{r, make_joint_with_species}
sepal_species <- iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width,
                       color = Species)) +
  geom_density_2d(size = 1.1) +
  geom_point(alpha = 0.5, size = 3) +
  ggthemes::scale_color_colorblind() +
  theme_bw() +
  theme(legend.position = "bottom")
```

Add in the marginal grouped density plots.  

```{r, add_marginal_species_dens}
sepal_species_marg <- ggMarginal(sepal_species, type = 'density', groupColour = TRUE )
```

Visualize the grouped joint and marginal densities.  

```{r, viz_species_density_groups}
print( sepal_species_marg )
```

We could also fill and color boxplots in the marginals if we would prefer.  

```{r, add_species_box_marg}
sepal_species_box <- ggMarginal( sepal_species, type = 'boxplot', 
                                 groupFill = TRUE, groupColour = TRUE,
                                 xparams = list(alpha = 0.35),
                                 yparams = list(alpha = 0.35))
```

Visualize.  

```{r, viz_species_box_marg}
print( sepal_species_box )
```

