---
title: "CMPINF 2130 Summer 2021 - Week 08"
subtitle: "reshaping data"
author: "Dr. Joseph P. Yurko"
date: "July 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries

```{r, load_tidyverse}
library(tidyverse)
```

## Anscombe

```{r, check_anscombe}
anscombe
```

The Anscombe data set in base `R` is **not** tidy. Each column name contains 2 pieces of information. We want to separate the column names such that each column is a single entity.  

We will need to reshape the data and then manipulate the data to create the tidy and easy to model data set.  
We must **melt** the data into a long format data set. Pandas refers to melting or making a data *molten* when we reshape wide to long. Currently in the `tidyverse` converting from wide to long format is performed with the `pivot_longer()` function.  

Before making the data long, I like to include a rowid column. The `cols` argument in `pivot_longer()` is how we instruct what columns "gather" or "make molten". We can use the `dplyr` select-like terminology to identify those columns.  

```{r, ans_1}
anscombe %>% 
  tibble::rowid_to_column() %>% 
  pivot_longer(cols = -rowid)
```

We can check that we still have the same number of observations per variable with `count()`.  

```{r, ans_2}
anscombe %>% 
  tibble::rowid_to_column() %>% 
  pivot_longer(cols = -rowid) %>% 
  count(name)
```

We can provide a character string to identify the columns with the `cols` argument.  

```{r, ans_3}
anscombe %>% 
  tibble::rowid_to_column() %>% 
  pivot_longer(cols = !c("rowid"), names_to = "variable_name", values_to = "variable_value")
```

The `tidyr` package has a helper function dedicated to separating character strings based on patterns or positions.  

```{r, ans_4}
anscombe %>% 
  tibble::rowid_to_column() %>% 
  pivot_longer(cols = !c("rowid")) %>% 
  tidyr::separate(name,
                  c("var_name", "dataset"),
                  sep = 1,
                  remove = FALSE)
```

As a check, the Anscombe data set consists of 4 sub-datasets or groups.  

```{r, ans_5}
anscombe %>% 
  tibble::rowid_to_column() %>% 
  pivot_longer(cols = !c("rowid")) %>% 
  tidyr::separate(name,
                  c("var_name", "dataset"),
                  sep = 1,
                  remove = TRUE) %>% 
  count(dataset, var_name)
```

We ultimately want to relate `y` and `x` together. We want to be able to fit linear models with a linear relationship between `y` and `x`.  

$$ 
\mathrm{trend_n} = \beta_0 + \beta_1 \times x_n
$$

The likelihood is a Gaussian:  

$$ 
y_n \mid \mathrm{trend_n}, \sigma \sim \mathrm{normal} \left( \mathrm{trend_n}, \sigma \right) 
$$

We need to convert the long format into a wide format again. In `tidyverse` this is accomplished via `pivot_wider()`. We must specify the column the new wider columns will be named "from" and the column the new wider values or entries come from.  

```{r, ans_6}
anscombe %>% 
  tibble::rowid_to_column() %>% 
  pivot_longer(cols = !c("rowid")) %>% 
  tidyr::separate(name,
                  c("var_name", "dataset"),
                  sep = 1,
                  remove = TRUE) %>% 
  pivot_wider(names_from = "var_name", values_from = "value")
```

Let's assign the tidy data set to an object.  

```{r, ans_7}
ans_tidy <- anscombe %>% 
  tibble::rowid_to_column() %>% 
  pivot_longer(cols = !c("rowid")) %>% 
  tidyr::separate(name,
                  c("var_name", "dataset"),
                  sep = 1,
                  remove = TRUE) %>% 
  pivot_wider(names_from = "var_name", values_from = "value")
```

As a check, let's look at the observations for the first data set.  

```{r, ans_1_check}
anscombe %>% 
  select(ends_with("1"))
```

Out tidy data set associated with Anscombe data set 1:  

```{r, ans_1_check_tidy}
ans_tidy %>% 
  filter(dataset == "1")
```

We can easily group and summarize the tidy data. Let's calculate summary statistics for `x` and `y` associated with each group.  

```{r, ans_8}
ans_tidy %>% 
  group_by(dataset) %>% 
  summarise(num_rows = n(),
            avg_x = mean(x),
            var_x = var(x),
            avg_y = mean(y),
            var_y = var(y)) %>% 
  ungroup()
```

The basic summary stats of sample averages and sample variances are essentially the same in all 4 Anscome groups. All 4 groups have the same linear relationship regression model! To show that, we will use nesting, functional programming with `purrr` and some help from the `broom` package.  

```{r, ans_9}
ans_tidy %>% 
  group_by(dataset) %>% 
  tidyr::nest() %>% 
  mutate(lm_mod = purrr::map(data, ~lm(formula = y ~ x, data = .)),
         lm_results = purrr::map(lm_mod, broom::glance),
         lm_coef = purrr::map(lm_mod, broom::tidy))
```

Let's extract coefficients and their standard errors for each for 4 fit models.  

```{r, ans_10}
ans_tidy %>% 
  group_by(dataset) %>% 
  tidyr::nest() %>% 
  mutate(lm_mod = purrr::map(data, ~lm(formula = y ~ x, data = .)),
         lm_results = purrr::map(lm_mod, broom::glance),
         lm_coef = purrr::map(lm_mod, broom::tidy)) %>% 
  select(dataset, lm_coef) %>% 
  tidyr::unnest(lm_coef)
```

Let's reshape into a wide format and look at the estimates for the intercept and slope across the 4 data sets.  

```{r, ans_11}
ans_tidy %>% 
  group_by(dataset) %>% 
  tidyr::nest() %>% 
  mutate(lm_mod = purrr::map(data, ~lm(formula = y ~ x, data = .)),
         lm_results = purrr::map(lm_mod, broom::glance),
         lm_coef = purrr::map(lm_mod, broom::tidy)) %>% 
  select(dataset, lm_coef) %>% 
  tidyr::unnest(lm_coef) %>% 
  select(dataset, term, estimate) %>% 
  pivot_wider(names_from = 'term', values_from = 'estimate')
```

Let's now look at the performance metrics.  

```{r, ans_12}
ans_tidy %>% 
  group_by(dataset) %>% 
  tidyr::nest() %>% 
  mutate(lm_mod = purrr::map(data, ~lm(formula = y ~ x, data = .)),
         lm_results = purrr::map(lm_mod, broom::glance),
         lm_coef = purrr::map(lm_mod, broom::tidy)) %>% 
  select(dataset, lm_results) %>% 
  tidyr::unnest(lm_results)
```

Let's now look at the data. We will use a scatter plot and facet by the data set.  

```{r, viz_ans_1}
ans_tidy %>% 
  ggplot(mapping = aes(x = x, y = y)) +
  geom_point(size = 2) +
  facet_wrap(~dataset, labeller = "label_both") +
  theme_bw()
```

We can include the linear relationship linear model on the facets that we had previously fit.  

```{r, viz_ans_2}
ans_tidy %>% 
  ggplot(mapping = aes(x = x, y = y)) +
  geom_point(size = 2) +
  geom_smooth(formula = y ~ x,
              method = lm) +
  facet_wrap(~dataset, labeller = "label_both") +
  theme_bw()
```

