---
title: "CMPINF 2130 Summer 2021 - Week 04"
author: "Dr. Joseph P. Yurko"
date: "6/7/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages

```{r, load_tidyverse_pkg}
library(tidyverse)
```

## Review scatter plot essentials

Let's use the `iris` data set again and make a scatter plot between the Sepal variables.  

**Start simple**  

```{r, iris_scatter_1}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point() +
  theme_bw()
```

We can add complexity by mapping additional variables to other aesthetics of the geometric object.  

```{r, iris_scatter_2}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(mapping = aes(color = Species,
                           shape = Species),
             size = 4.5, alpha = 0.5) +
  ggthemes::scale_color_colorblind() +
  theme_bw()
```

Coloring by continuous variables.  

```{r, iris_scatter_3}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(mapping = aes(color = Petal.Length),
             size = 4.5, alpha = 0.5) +
  scale_color_viridis_c() +
  theme_bw()
```

Add further complexity through facets.  

```{r, iris_scatter_4}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(mapping = aes(color = Petal.Length),
             size = 4.5, alpha = 0.5) +
  facet_wrap( ~ Species ) +
  scale_color_viridis_c() +
  theme_bw() +
  theme(legend.position = "top")
```

Create a binary version of `Petal.Width` and map that binary variable to the marker shape.  

```{r, iris_scatter_5}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(mapping = aes(color = Petal.Length,
                           shape = Petal.Width > median(Petal.Width)),
             size = 4.5, alpha = 0.5) +
  facet_wrap( ~ Species ) +
  scale_color_viridis_c() +
  theme_bw() +
  theme(legend.position = "top")
```

We can keep Petal.Width continuous by mapping it to the marker size and the marker transparency.  

```{r, iris_scatter_6}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(mapping = aes(color = Petal.Length,
                           size = Petal.Width,
                           alpha = Petal.Width)) +
  facet_wrap( ~ Species ) +
  scale_color_viridis_c() +
  theme_bw() +
  theme(legend.position = "top")
```

## Lines, paths, and trends

A line shows continuity in the x-axis ordering.  

```{r, iris_line_1}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_line() +
  geom_point(size = 2.5) +
  theme_bw()
```

The `geom_line()` is particularly useful when we want to show the **order** of a value with respect to the x-axis. A "RUN PLOT" or "RUN CHART" is one type of a figure where `geom_line()` is the kind of geom to use.  

```{r, iris_line_2}
iris %>% 
  tibble::rowid_to_column() %>% 
  ggplot(mapping = aes(x = rowid, y = Sepal.Width)) +
  geom_line() +
  geom_point(size = 2.5) +
  theme_bw()
```


Trends let us see the "signal through the noise". We will focus at first on linear relationships between the `y` and `x` aesthetics.  

First, remake the scatter plot where we color by Species.  

```{r, iris_trend_1}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(mapping = aes(color = Species,
                           shape = Species),
             size = 4.5, alpha = 0.5) +
  ggthemes::scale_color_colorblind() +
  theme_bw()
```


Add in a trend line to this plot with `geom_smooth()`. Force a linear model with `method = lm`.  

```{r, iris_trend_2}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(mapping = aes(color = Species,
                           shape = Species),
             size = 4.5, alpha = 0.5) +
  geom_smooth(method = lm) +
  ggthemes::scale_color_colorblind() +
  theme_bw()
```

Let's specify the formula we want so the warning is not displayed.  

```{r, iris_trend_3}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(mapping = aes(color = Species,
                           shape = Species),
             size = 4.5, alpha = 0.5) +
  geom_smooth(method = lm,
              formula = y ~ x) +
  ggthemes::scale_color_colorblind() +
  theme_bw()
```

Last week we looked at the correlation coefficients.  

```{r, iris_corr_1}
iris %>% 
  select(starts_with("Sepal.")) %>% 
  cor()
```

But we know that's not the case when we consider grouping by the Species.  

```{r, iris_corr_2}
iris %>% 
  select(starts_with("Sepal."), Species) %>% 
  group_by(Species) %>% 
  group_modify( ~ as.data.frame(cor(.x)))
```


Let's now map Species to the color aesthetic in `geom_smooth()`.  

```{r, iris_trend_4}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(mapping = aes(color = Species,
                           shape = Species),
             size = 4.5, alpha = 0.5) +
  geom_smooth(method = lm,
              formula = y ~ x,
              mapping = aes(color = Species)) +
  ggthemes::scale_color_colorblind() +
  theme_bw()
```

We could have got the same result if we mapped `Species` to `color` in the parent call.  

```{r, iris_trend_5}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width,
                       color = Species)) +
  geom_point(mapping = aes(shape = Species),
             size = 4.5, alpha = 0.5) +
  geom_smooth(method = lm,
              formula = y ~ x) +
  ggthemes::scale_color_colorblind() +
  theme_bw()
```


If we do not map color in the parent `ggplot()` call, we can have separate smoothing lines for "all data" and for the individual Species.  

```{r, iris_trend_6}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(mapping = aes(color = Species,
                           shape = Species),
             size = 4.5, alpha = 0.5) +
  # smooth based on all data
  geom_smooth(method = lm, formula = y ~ x,
              color = 'red', fill = 'red') +
  # smooth per group 
  geom_smooth(method = lm, formula = y ~ x,
              mapping = aes(color = Species)) +
  ggthemes::scale_color_colorblind() +
  theme_bw()
```

We can remove the points, by simply **not** including the `geom_point()` layer.  

**I DO NOT RECOMMEND THIS TYPE OF FIGURE**  

```{r, iris_trend_7}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  # smooth based on all data
  geom_smooth(method = lm, formula = y ~ x,
              color = 'red', fill = 'red') +
  # smooth per group 
  geom_smooth(method = lm, formula = y ~ x,
              mapping = aes(color = Species)) +
  ggthemes::scale_color_colorblind() +
  theme_bw()
```

For right now, I'll remove the confidence interval just so we can focus on the trend line.  

**BUT I DON'T LIKE TO NORMALLY!!!!!**  

```{r, iris_trend_8}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  # smooth based on all data
  geom_smooth(method = lm, formula = y ~ x,
              color = 'red', fill = 'red',
              se = FALSE) +
  # smooth per group 
  geom_smooth(method = lm, formula = y ~ x,
              mapping = aes(color = Species),
              se = FALSE) +
  ggthemes::scale_color_colorblind() +
  theme_bw()
```

Lines have similiar aesthetics as points.  

```{r, iris_trend_9}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  # smooth based on all data
  geom_smooth(method = lm, formula = y ~ x,
              color = 'red', fill = 'red',
              se = FALSE,
              size = 2.5) +
  # smooth per group 
  geom_smooth(method = lm, formula = y ~ x,
              mapping = aes(color = Species),
              se = FALSE,
              size = 1.25) +
  ggthemes::scale_color_colorblind() +
  theme_bw()
```

linetype is analogous to marker shape.  

```{r, iris_trend_10}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  # smooth based on all data
  geom_smooth(method = lm, formula = y ~ x,
              color = 'red', fill = 'red',
              se = FALSE,
              size = 2.5) +
  # smooth per group 
  geom_smooth(method = lm, formula = y ~ x,
              mapping = aes(color = Species,
                            linetype = Species),
              se = FALSE,
              size = 1.25) +
  ggthemes::scale_color_colorblind() +
  theme_bw()
```

We have up to now used `color` to force a grouping of the data "behind the scenes". We can be more explicit and state that the `group` aesthetic is `Species`.  

```{r, iris_trend_11}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(mapping = aes(color = Species,
                           shape = Species),
             size = 4.5, alpha = 0.5) +
  # smooth based on all data
  geom_smooth(method = lm, formula = y ~ x,
              color = 'purple',
              mapping = aes(group = 1),
              se = FALSE,
              size = 2.5, linetype = 'dashed') +
  # smooth per group 
  geom_smooth(method = lm, formula = y ~ x,
              mapping = aes(group = Species),
              color = 'forestgreen', size = 2,
              se = FALSE) +
  ggthemes::scale_color_colorblind() +
  theme_bw()
```

## Statistical transformations

Statistical transformation are ways of **deriving** or calculating aesthetics from "raw" data.  

Essentially, the variable we will plot on the graph does **not** exist in the raw data set.  

Linear models or *smoothers* are one type of statistical transformation, but the simplest one is to **count**.  

How many rows exist for each *level* of `Species`.  

```{r, iris_count_1}
iris %>% count(Species)
```

Behind the scenes `count()` must first `group_by()` and then `summarize()` the data. Split-Apply-Combine: split the data into separate groups, apply a *summary* function, and combine the data back together.  

```{r, iris_count_2}
iris %>% 
  group_by(Species) %>% 
  summarise(num_rows = n())
```

Can check that this is correct via filtering.  

```{r, iris_count_3}
iris %>% 
  filter(Species == 'setosa') %>% 
  nrow()
```

```{r, iris_count_4}
iris %>% 
  filter(Species == 'versicolor') %>% 
  nrow()
```

```{r, iris_count_4_b}
iris %>% 
  group_by(Species) %>% 
  tidyr::nest() %>% 
  mutate(num_rows = purrr::map_dbl(data, nrow))
```

A bar graph or bar chart is useful to display counts, the taller the bar the more counts!  

```{r, iris_bar_1}
iris %>% 
  ggplot(mapping = aes(x = Species)) +
  geom_bar() +
  theme_bw()
```

Let's make the bar chart ourselves rather than relying on the the counting behind the scenes. To use `geom_bar()` after grouping and summarizing, therefore having the counts per level we must force `stat = 'identity'`.  

```{r, iris_bar_2}
iris %>% 
  count(Species) %>% 
  ggplot(mapping = aes(x = Species, y = n)) +
  geom_bar(stat = 'identity') +
  theme_bw()
```

Alternatively, we can use the `geom_col()` which does **not** perform statistical transformations.  

```{r, iris_bar_3}
iris %>% 
  count(Species) %>% 
  ggplot(mapping = aes(x = Species, y = n)) +
  geom_col() +
  theme_bw()
```

One last way to create a bar chart is with `stat_count()` directly where we force the `geom` to be `'bar'`.  

```{r, iris_bar_4}
iris %>% 
  ggplot(mapping = aes(x = Species)) +
  stat_count(geom = 'bar') +
  theme_bw()
```

Include the count or the number of rows as text on top of the bars. We can instruct `geom_text()` to perform a statistical transformation and thus COUNT the number of rows.  

```{r, iris_bar_5}
iris %>% 
  ggplot(mapping = aes(x = Species)) +
  geom_bar() +
  geom_text(stat = 'count',
            mapping = aes(label = stat(count))) +
  theme_bw()
```

We can manually "nudge" the position of the text.  

```{r, iris_bar_6}
iris %>% 
  ggplot(mapping = aes(x = Species)) +
  geom_bar() +
  geom_text(stat = 'count',
            mapping = aes(label = stat(count)),
            nudge_y = 2.5, color = 'red', size = 7.5) +
  theme_bw()
```

We could "nudge" below the bar.  

```{r, iris_bar_7}
iris %>% 
  ggplot(mapping = aes(x = Species)) +
  geom_bar() +
  geom_text(stat = 'count',
            mapping = aes(label = stat(count)),
            nudge_y = -2.5, color = 'white', size = 7.5) +
  theme_bw()
```

We can also map Species to the fill of the bar.  

```{r, iris_bar_8}
iris %>% 
  ggplot(mapping = aes(x = Species)) +
  geom_bar(mapping = aes(fill = Species)) +
  geom_text(stat = 'count',
            mapping = aes(label = stat(count)),
            nudge_y = 2.5, color = 'red', size = 7.5) +
  ggthemes::scale_fill_colorblind() +
  theme_bw()
```

We can turn off the legend since the x-axis is the fill.  

```{r, iris_bar_9}
iris %>% 
  ggplot(mapping = aes(x = Species)) +
  geom_bar(mapping = aes(fill = Species)) +
  geom_text(stat = 'count',
            mapping = aes(label = stat(count)),
            nudge_y = 2.5, color = 'red', size = 7.5) +
  ggthemes::scale_fill_colorblind(guide=FALSE) +
  theme_bw()
```

Difference between fill and color.  

```{r, iris_bar_10}
iris %>% 
  ggplot(mapping = aes(x = Species)) +
  geom_bar(mapping = aes(color = Species), size = 5.5) +
  geom_text(stat = 'count',
            mapping = aes(label = stat(count)),
            nudge_y = 2.5, color = 'red', size = 7.5) +
  ggthemes::scale_color_colorblind(guide=FALSE) +
  theme_bw()
```

Let's create a binary variable based on the value `Sepal.Length` relative to the average `Sepal.Length`.  

```{r, calc_length_avg_1}
mean(iris$Sepal.Length)
```

```{r, calc_length_avg_2}
iris %>% summarise(mean(Sepal.Length))
```

Check the number of rows where `Sepal.Length` is greater than the average value.  

```{r, check_num_rows_length}
iris %>% 
  filter(Sepal.Length > mean(Sepal.Length)) %>% 
  nrow()
```

Let's visualize this as a bar chart.  

```{r, iris_length_bar_1}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length > mean(Sepal.Length))) +
  geom_bar() +
  theme_bw()
```

How many rows per group have the `Sepal.Length` greater than the overall average?  

```{r, iris_length_groups_check}
iris %>% 
  mutate(length_bin = Sepal.Length > mean(Sepal.Length)) %>% 
  count(Species, length_bin)
```

Visualize the counts with a bar chart.  

```{r, iris_bar_11}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length > mean(Sepal.Length))) +
  geom_bar(mapping = aes(fill = Species)) +
  ggthemes::scale_fill_colorblind() +
  theme_bw()
```


I prefer to rearrange the fill into **separate** and **distinct** bars that are next to each other. In `ggplot2` this is known as **dodging** the position.  

```{r, iris_bar_12}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length > mean(Sepal.Length))) +
  geom_bar(mapping = aes(fill = Species),
           position = "dodge") +
  ggthemes::scale_fill_colorblind() +
  theme_bw()
```


We can also include the text for the counts. The `geom_text()` does **NOT** know the grouping structure.  

```{r, iris_bar_13}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length > mean(Sepal.Length))) +
  geom_bar(mapping = aes(fill = Species),
           position = "dodge") +
  geom_text(stat = 'count',
            mapping = aes(label = stat(count))) +
  ggthemes::scale_fill_colorblind() +
  theme_bw()
```

We can override or instruct how to group the text via the `group` aesthetic.  

```{r, iris_bar_14}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length > mean(Sepal.Length))) +
  geom_bar(mapping = aes(fill = Species),
           position = "dodge") +
  geom_text(stat = 'count',
            mapping = aes(group = Species,
                          label = stat(count))) +
  ggthemes::scale_fill_colorblind() +
  theme_bw()
```

Adjusting the text position is painful.  

```{r, iris_bar_15}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length > mean(Sepal.Length))) +
  geom_bar(mapping = aes(fill = Species),
           position = "dodge") +
  geom_text(stat = 'count',
            mapping = aes(group = Species,
                          label = stat(count)),
            position = position_dodge(0.75),
            color = 'red', size = 7) +
  ggthemes::scale_fill_colorblind() +
  theme_bw()
```


Let's set the x-axis back to Species and the fill to the conditional test.  

```{r, iris_bar_16}
iris %>% 
  ggplot(mapping = aes(x = Species)) +
  geom_bar(mapping = aes(fill = Sepal.Length > mean(Sepal.Length)),
           position = "dodge") +
  ggthemes::scale_fill_calc() +
  theme_bw() +
  theme(legend.position = "top")
```

Essentially we are using the bar fill to denote the number of rows per **combination**. Combinations can also be displayed via heat maps.  

A heat map is essentially a scatter plot. It consists of x position and y position BUT instead of points we will use a TILE.  

```{r, iris_heatmap_1}
iris %>% 
  count(Species, Sepal.Length > mean(Sepal.Length)) %>% 
  ggplot(mapping = aes(x = `Sepal.Length > mean(Sepal.Length)`,
                       y = Species)) +
  geom_tile(color = 'black', fill = 'grey') +
  theme_bw()
```

Map the number of observations to the fill of the tile.  

```{r, iris_heatmap_2}
iris %>% 
  count(Species, Sepal.Length > mean(Sepal.Length)) %>% 
  ggplot(mapping = aes(x = `Sepal.Length > mean(Sepal.Length)`,
                       y = Species)) +
  geom_tile(color = 'black', 
            mapping = aes(fill = n)) +
  scale_fill_viridis_c() +
  theme_bw()
```

Add text to show the number of rows by mapping the `geom_text()` label to `n`.  

```{r, iris_heatmap_3}
iris %>% 
  count(Species, Sepal.Length > mean(Sepal.Length)) %>% 
  ggplot(mapping = aes(x = `Sepal.Length > mean(Sepal.Length)`,
                       y = Species)) +
  geom_tile(color = 'black', 
            mapping = aes(fill = n)) +
  geom_text(mapping = aes(label = n), size = 8) +
  scale_fill_viridis_c() +
  theme_bw()
```

Lastly, map the text color to change with the number of rows.  

```{r, iris_heatmap_4}
iris %>% 
  count(Species, Sepal.Length > mean(Sepal.Length)) %>% 
  ggplot(mapping = aes(x = `Sepal.Length > mean(Sepal.Length)`,
                       y = Species)) +
  geom_tile(color = 'black', 
            mapping = aes(fill = n)) +
  geom_text(mapping = aes(label = n,
                          color = n > median(n)), 
            size = 8) +
  scale_fill_viridis_c(guide = FALSE) +
  scale_color_manual(guide = FALSE,
                     values = c("TRUE" = 'black',
                                "FALSE" = "white")) +
  theme_bw()
```


We could show this information with points and sizes of points.  

```{r, iris_count_map_1}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length > mean(Sepal.Length),
                       y = Species)) +
  geom_count() +
  theme_bw()
```


