---
title: "CMPINF 2130 Summer 2021 - Week 02"
subtitle: "R Syntax and data manipulation"
author: "Dr. Joseph P. Yurko"
date: "5/24/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Vector review

Let's create a simple vector with values from 1 to 11.  

```{r, vector_example_1}
x_vec <- 1:11
```

Multiply two vectors together. We created one of them but the second has not been made. If this code chunk is executed things break.  

```{r, eval=FALSE}
x.vec * x_vec
```

Display the object to the screen.  

We need two spaces after a line to tell Markdown to create a new "paragraph".  

```{r, vector_example_2}
x_vec
```

Make another vector.  

```{r, vector_example_3}
x.vec <- 1:11
```

We can subset a vector by applying a conditional test.  

```{r, vector_example_4}
x_vec > 5
```

We can pass in the boolean vector to SUBSET a vector.  

```{r, vector_example_5}
x_vec[ x_vec > 5 ]
```

If we wanted to subset by position or index, we can pass in the integer index value.  

```{r, vector_example_6}
x_vec[ 1 ]
```

```{r, vector_example_7}
x_vec[ 1:4 ]
```

We can also apply equality conditions.  

```{r, vector_example_8}
x_vec == 5
```

```{r, vector_example_9}
x_vec[ x_vec == 5 ]
```

We can also have an **OR** statement. For example, lets use does the element equal 2, 3, or 7.  

```{r, vector_example_10}
x_vec[ x_vec %in% c(2, 3, 7) ]
```

We can have the **NOT IN** operation.  

```{r, vector_example_11}
x_vec[ !x_vec %in% c(2, 3, 7) ]
```

R vectors homogeneous. They can only store a single data type.  

## Lists

It is a data container. It's a blend of the Python list and a dictionary.  

```{r, list_example_1}
my_list <- list(a = 1:5,
                b = as.character(1:3),
                c = letters[1:7],
                d = c(TRUE, FALSE, TRUE, FALSE))
```

```{r, list_example_2}
my_list
```

We can use the `$` operator to access the fields in a list.  

```{r, list_example_3}
my_list$a

my_list$d
```

We can also access the elements or fields of a list with double bracket notation.  

```{r, list_example_4}
my_list[[1]]
```

```{r, list_example_5}
my_list[[4]]
```

We can use character strings with the double bracket notation.  

```{r, list_example_6}
my_list[['a']]

my_list[['d']]
```

Use a for loop to iterate over the elements of the list and print the length of each field (object in a list) to the screen.  

```{r, list_example_7}
for(n in (1:length(my_list)) )
{
  print( length( my_list[[n]] ) )
}
```

Instead I prefer the `seq_along()` function to create a vector to iterate over.  

```{r, list_example_8}
seq_along(my_list)
```

```{r, list_examplee_9}
for(n in seq_along(my_list) )
{
  print( length( my_list[[n]] ) )
}
```

We can also iterate over the element or field names of the list.  

```{r, list_example_10}
names( my_list )
```

```{r, list_example_11}
for(a_name in names(my_list) )
{
  print( length( my_list[[a_name]] ) )
}
```

## Data frames

A `data.frame` is a specialized list. Each list contained in a `data.frame` must store the same data type, but different lists can store different data types. Every list in a `data.frame` MUST have the same length.  

```{r, df_example_1}
my_df <- data.frame(x1 = 1:4, 
                    x2 = c('yes', 'no', 'maybe', 'hello'), 
                    x3 = seq(-1, 1, length.out = 4), 
                    x4 = c(TRUE, FALSE, FALSE, TRUE))
```

```{r, df_example_2}
my_df
```

Because a `data.frame` is just a specialized list, we can access columns with the `$` operator.  

```{r, df_example_3}
my_df$x2
```

When we use the `$` operator to access or *select* a column from the `data.frame` it returns a regular vector.  

```{r, df_example_4}
class( my_df$x2 )
```

We can also use the column index to select a column.  

```{r, df_example_5}
my_df[1, 1]
```

```{r, df_example_6}
my_df[ , 1]
```

To prevent a `data.frame` from dropping down to a vector we must use the `drop = FALSE` argument.  

```{r, df_example_7}
my_df[, 1, drop=FALSE]
```

Select columns with character strings for the names.  

```{r, df_example_8}
my_df[, "x1", drop=FALSE]
```

We can also select multiple columns.  

```{r, df_example_9}
my_df[, c("x1", "x2"), drop=FALSE]
```

## Tidyverse

Load the tidyverse.  

```{r, load_tidyverse_pkg}
library(tidyverse)
```

The RStudio cheat sheets are good resource for the tidyverse family of packages.  

[RStudio Cheat sheet link](https://www.rstudio.com/resources/cheatsheets/)  

We will manipulate the `data.frame` with the `dplyr` package.  

```{r, dplyr_example_1}
select(my_df, x1)
```

Select two columns.  

```{r, dplyr_example_2}
select(my_df, x1, x2)
```

`filter()` is for the applying conditional subsets for rows.  

```{r, dplyr_example_3}
filter(my_df, x3 > 0)
```

The `dplyr` package also loads in the `%>%` forward-pipe operator from the `magrittr` package.  

```{r, pipe_example_1}
my_df %>% select(x1, x2)
```

We can string or chain together data manipulation steps into a pipeline.  

```{r, pipe_example_2}
my_df %>% 
  filter(x3 > 0) %>% 
  select(x1, x2)
```

```{r, pipe_example_3}
select( filter( my_df, x3 > 0), x1, x2 )
```

Add a column we use `mutate()`.  

```{r, pipe_example_4}
my_df %>% 
  mutate(x5 = x1 * x3)
```


```{r}
my_df
```

```{r}
my_df %>% 
  mutate(x5 = x1 * x3) %>% 
  filter(abs(x5) > 1) %>% 
  select(-x5)
```

