---
title: "CMPINF 2130 Summer 2021 - Week 02"
subtitle: "Intro to ggplot2"
author: "Dr. Joseph P. Yurko"
date: "5/24/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages

```{r, load_tidyverse_pkg}
library(tidyverse)
```

## Anscombe data set

```{r, check_dataset}
anscombe %>% class()
```

```{r, check_ans_str}
anscombe %>% str()
```

Or the tidyverse way.  

```{r, check_ans_glimpse}
anscombe %>% glimpse()
```

```{r, asn_check_head}
anscombe %>% head()
```

Select the `x1` and `y1` columns and rename them `x` and `y`.  

```{r, subset_ans}
a1 <- anscombe %>% 
  # select(x = x1, y = y1) # or could have done this
  dplyr::select(ends_with("1")) %>% 
  rename(x = x1, y = y1)
```

```{r, check_a1_glimpse}
a1 %>% glimpse()
```


## ggplot2

A `ggplot2` graphic starts with a call to the `ggplot()` function. I refer to the `ggplot()` call as the "parent ggplot call".  

```{r, gg_example_1}
ggplot()
```

The first argument to `ggplot()` function is `data`.  

```{r, gg_example_2}
a1 %>% 
  ggplot()
```

Map the basic aesthetics of a figure the `x` and `y` aesthetics (the x and y axes variables).  

```{r, gg_example_3}
a1 %>% 
  ggplot(mapping = aes(x = x, y = y))
```


Check the data summary.  

```{r, check_a1_summary}
a1 %>% summary()
```

How to visualize the aesthetics.  Geometric objects are defined by the suite of `geom_*` functions from `ggplot2`. A scatter plot is represented by the `geom_point()` geometric object.  

```{r, gg_example_4}
a1 %>% 
  ggplot(mapping = aes(x = x, y = y)) +
  geom_point()
```


Change the marker color from the default to `'steelblue'`. Do we specify the `color` argument inside `aes()` or outside `aes()`? Let's see what happens if we set `color` inside `aes()`.  

```{r, gg_example_5}
a1 %>% 
  ggplot(mapping = aes(x = x, y = y)) +
  geom_point(mapping = aes(color = 'steelblue'))
```


```{r, gg_example_5b}
a1 %>% 
  ggplot(mapping = aes(x = x, y = y)) +
  geom_point(mapping = aes(color = 'steelblue')) +
  ggthemes::scale_color_colorblind()
```

```{r, gg_example_5c}
a1 %>% 
  ggplot(mapping = aes(x = x, y = y)) +
  geom_point(mapping = aes(color = 'steelblue')) +
  ggthemes::scale_color_calc()
```

```{r}
a1 %>% 
  ggplot(mapping = aes(x = x, y = y)) +
  geom_point(mapping = aes(color = 'a value I made up')) +
  ggthemes::scale_color_calc()
```



`ggplot()` thinks there's a variable with a value `steelblue` and the legend exists to reference the possible values represented by the color.  

If we want to "hard code" a visual quality a visual aesthetic we place that assignment **outside** the `aes()` function.  

```{r, gg_example_6}
a1 %>% 
  ggplot(mapping = aes(x = x, y = y)) +
  geom_point(color = 'steelblue')
```

Let's change the marker size.  

```{r, gg_example_7}
a1 %>% 
  ggplot(mapping = aes(x = x, y = y)) +
  geom_point(color = 'steelblue', size = 5.5)
```

Change the marker shape with the `shape` argument.  

```{r, gg_example_8}
a1 %>% 
  ggplot(mapping = aes(x = x, y = y)) +
  geom_point(color = 'steelblue', size = 5.5, shape = 15)
```

## Iris

```{r, iris_1}
iris %>% glimpse()
```

```{r, iris_2}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(mapping = aes(color = Species,
                           shape = Species,
                           size = Species),
             alpha = 0.5)
```

Let's use a manual size scale for the Species.  

```{r, iris_3}
my_size_values <- c(1, 2, 4)

names(my_size_values) <- iris %>% distinct(Species) %>% pull()

iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(mapping = aes(color = Species,
                           shape = Species,
                           size = Species),
             alpha = 0.5) +
  scale_size_manual(values = my_size_values)
```

```{r, iris_4}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(mapping = aes(color = Species,
                           shape = Species,
                           size = Species),
             alpha = 0.5) +
  scale_size_manual(values = my_size_values) +
  ggthemes::scale_color_colorblind()
```

```{r, iris_5}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(mapping = aes(color = Species,
                           shape = Species,
                           size = Species),
             alpha = 0.5) +
  scale_size_manual(values = my_size_values) +
  ggthemes::scale_color_calc()
```

If you're an Excel user.  

```{r, iris_6}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(mapping = aes(color = Species,
                           shape = Species,
                           size = Species),
             alpha = 0.5) +
  scale_size_manual(values = my_size_values) +
  ggthemes::scale_color_excel() +
  theme_classic()
```

