---
title: "CMPINF 2130 Summer 2021 - Week 03"
subtitle: "Points, text, lines, and paths"
author: "Dr. Joseph P. Yurko"
date: "6/1/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages

```{r, load_tidyverse}
library(tidyverse)
```

## Anscombe data

```{r, glimpse_anscombe}
anscombe %>% glimpse()
```

The `anscombe` data are a wide object.  

```{r, check_ans_wide}
anscombe
```

Last week we focused on the first data set, so the `x` and `y` variables that end in `"1"`. Let's continue to focus on the first data set today.  

```{r, separate_first_data}
a1 <- anscombe %>% 
  select(ends_with("1")) %>% 
  rename(x = x1, y = y1)

a1 %>% glimpse()
```

Visualizing the relationship between `x` and `y` as a scatter plot.  

```{r, ans_a1_viz}
a1 %>% 
  ggplot(mapping = aes(x = x, y = y)) +
  geom_point(size = 5.5) +
  theme_bw()
```

Want to display text at specific points in the plot. It's useful supplemental information. A simple type of supplemental information to display is the row order index on a point. Our data set does not include such an index.  

```{r, check_a1_rows}
a1
```

We could manually add in a column for the row index, but the `tibble` package which is part of tidyverse has the `rowid_to_column()` to help out. 

```{r, a1_add_rowid}
a1 %>% 
  tibble::rowid_to_column()
```

Include the rowid as text on the plot. The text is a geometric object. It has aesthetics, such as the `x` and `y` coordinate positions. And the `label` which is what we want to display at the position. There are two ways of showing text objects, the `geom_label()` and `geom_text()`.  

```{r, a1_viz_label}
a1 %>% 
  tibble::rowid_to_column() %>% 
  ggplot(mapping = aes(x = x, y = y)) +
  geom_label(mapping = aes(label = rowid)) +
  theme_bw()
```

If we change the theme to a dark them the label "bubble" by default is still white color.  

```{r, a1_viz_label_2}
a1 %>% 
  tibble::rowid_to_column() %>% 
  ggplot(mapping = aes(x = x, y = y)) +
  geom_label(mapping = aes(label = rowid)) +
  theme_dark()
```


`geom_text()` is more streamlined, it only shows the text label and does not include the "bubble" around the text.  

```{r, a1_viz_text_1}
a1 %>% 
  tibble::rowid_to_column() %>% 
  ggplot(mapping = aes(x = x, y = y)) +
  geom_text(mapping = aes(label = rowid)) +
  theme_bw()
```

We can modify the size of the text with the `size` aesthetic.  

```{r, a1_viz_text_2}
a1 %>% 
  tibble::rowid_to_column() %>% 
  ggplot(mapping = aes(x = x, y = y)) +
  geom_text(mapping = aes(label = rowid),
            size = 8.5) +
  theme_bw()
```


We can also change the color.  

```{r, a1_viz_text_3}
a1 %>% 
  tibble::rowid_to_column() %>% 
  ggplot(mapping = aes(x = x, y = y)) +
  geom_text(mapping = aes(label = rowid),
            size = 8.5, color = 'red') +
  theme_bw()
```

We can combine multiple geometric objects together to creat a multi layered graphic. We add layers together to "build up" the final graphic.  

```{r, a1_viz_text_4}
a1 %>% 
  tibble::rowid_to_column() %>% 
  ggplot(mapping = aes(x = x, y = y)) +
  geom_point() +
  geom_text(mapping = aes(label = rowid)) +
  theme_bw()
```

Let's modify the points so that we have larger points of a darker color with text of a lighter color on top.  

```{r, a1_viz_text_5}
a1 %>% 
  tibble::rowid_to_column() %>% 
  ggplot(mapping = aes(x = x, y = y)) +
  geom_point(size = 7.5, color = 'black') +
  geom_text(mapping = aes(label = rowid),
            color = 'white') +
  theme_bw()
```

Text aesthetics can be mapped to variables in the data set. So let's try the text color and size mapped to the rowid.  

```{r, a1_viz_text_6}
a1 %>% 
  tibble::rowid_to_column() %>% 
  ggplot(mapping = aes(x = x, y = y)) +
  geom_point(size = 10, color = 'grey') +
  geom_text(mapping = aes(label = rowid,
                          color = rowid,
                          size = rowid)) +
  scale_color_viridis_c() +
  scale_size_continuous() +
  theme_bw()
```

So let's use a fill aesthetic for the markers and a color aesthetic for the font color. Specify a marker shape that has both fill and color stroke.  

```{r, a1_viz_text_7}
a1 %>% 
  tibble::rowid_to_column() %>% 
  ggplot(mapping = aes(x = x, y = y)) +
  geom_point(size = 10,
             mapping = aes(fill = rowid),
             shape = 21, color = 'black', stroke = 1.) +
  geom_text(mapping = aes(label = rowid,
                          color = rowid > 6)) +
  scale_fill_viridis_c() +
  scale_color_manual(values = c("TRUE" = 'black',
                                "FALSE" = 'white')) +
  theme_bw()
```

Turn off the legend guide for the text color and the marker fill.  

```{r, a1_viz_text_8}
a1 %>% 
  tibble::rowid_to_column() %>% 
  ggplot(mapping = aes(x = x, y = y)) +
  geom_point(size = 10,
             mapping = aes(fill = rowid),
             shape = 21, color = 'black', stroke = 1.) +
  geom_text(mapping = aes(label = rowid,
                          color = rowid > 6)) +
  scale_fill_viridis_c(guide = FALSE) +
  scale_color_manual(guide = FALSE,
                     values = c("TRUE" = 'black',
                                "FALSE" = 'white')) +
  theme_bw()
```

## Lines and paths

When we want to express continuity we want to use lines. The `geom_line()` lets us express continuity in the x-axis variable. So `geom_line()` connects **sequential** points based on the value of the `x` aesthetic. For small data sets, I like to combine `geom_line()` and `geom_point()` so I can see the exact data points getting connected. To see why let's first just look at the `geom_line()`.  

```{r, a1_viz_line_1}
a1 %>% 
  ggplot(mapping = aes(x = x, y = y)) +
  geom_line() +
  theme_bw()
```

Combining the line with the points lets us see that the "kinks" or the "mountain tops" are associated with data points.  

```{r, a1_viz_line_2}
a1 %>% 
  ggplot(mapping = aes(x = x, y = y)) +
  geom_line() +
  geom_point() +
  theme_bw()
```

The line is choppy because there's scatter in the data. Lines have aesthetics size is the line thickness, `linetype` controls if the line is solid, dashed, dotted, etc...  

Let's now combine the point, the text, and the line in our graphic.  

```{r, a1_viz_line_3}
a1 %>% 
  tibble::rowid_to_column() %>% 
  ggplot(mapping = aes(x = x, y = y)) +
  geom_line(size = 1.) +
  geom_point(size = 10,
             mapping = aes(fill = rowid),
             shape = 21, color = 'black', stroke = 1.) +
  geom_text(mapping = aes(label = rowid,
                          color = rowid > 6)) +
  scale_fill_viridis_c(guide = FALSE) +
  scale_color_manual(guide = FALSE,
                     values = c("TRUE" = 'black',
                                "FALSE" = 'white')) +
  theme_bw()
```

There is another way to express continuity...we can consider the row order in the data set. The `geom_path()` object connects points as the appear in the data.  

```{r, a1_viz_path_1}
a1 %>% 
  ggplot(mapping = aes(x = x, y = y)) +
  geom_path() +
  theme_bw()
```

Why is it all over the place? Because the "path" follows the sequential ROW ORDER **not** the sequential x-axis ordering.  

```{r, a1_viz_path_2}
a1 %>% 
  tibble::rowid_to_column() %>% 
  ggplot(mapping = aes(x = x, y = y)) +
  geom_path(size = 1.55) +
  geom_point(size = 10,
             mapping = aes(fill = rowid),
             shape = 21, color = 'black', stroke = 1.) +
  geom_text(mapping = aes(label = rowid,
                          color = rowid > 6)) +
  scale_fill_viridis_c(guide = FALSE) +
  scale_color_manual(guide = FALSE,
                     values = c("TRUE" = 'black',
                                "FALSE" = 'white')) +
  theme_bw()
```

Combine the `geom_line()` and the `geom_path()` a thick transparent for the path and a red dashed line for the sequential x-axis line.  

```{r, a1_viz_path_3}
a1 %>% 
  tibble::rowid_to_column() %>% 
  ggplot(mapping = aes(x = x, y = y)) +
  geom_path(size = 4.5, color = 'black', alpha = 0.25) +
  geom_line(size = 1.25, color = 'red', linetype = 'dashed') +
  geom_point(size = 10,
             mapping = aes(fill = rowid),
             shape = 21, color = 'black', stroke = 1.) +
  geom_text(mapping = aes(label = rowid,
                          color = rowid > 6)) +
  scale_fill_viridis_c(guide = FALSE) +
  scale_color_manual(guide = FALSE,
                     values = c("TRUE" = 'black',
                                "FALSE" = 'white')) +
  theme_bw()
```

The "best fit line" or regression line requires a statistical transformation and is considered its own geometric object. In ggplot2 it's known as the `geom_smooth()`. One way to "smooth" is with a linear trend. By default the `geom_smooth()` allows for a non-linear smoother, and so to force a linear relationship which want to specify `method = lm` and the `formula = y ~ x`.  

```{r, a1_trend_1}
a1 %>% 
  ggplot(mapping = aes(x = x, y = y)) +
  geom_point(size = 5.5) +
  geom_smooth(method = lm) +
  theme_bw()
```

Specify the formula to remove the message.  

```{r, a1_trend_2}
a1 %>% 
  ggplot(mapping = aes(x = x, y = y)) +
  geom_point(size = 5.5) +
  geom_smooth(formula = y ~ x,
              method = lm) +
  theme_bw()
```

To turn off the confidence interval set `se = FALSE`.  

```{r, a1_trend_3}
a1 %>% 
  ggplot(mapping = aes(x = x, y = y)) +
  geom_point(size = 5.5) +
  geom_smooth(formula = y ~ x,
              method = lm,
              se = FALSE) +
  theme_bw()
```

Combine the points, text, line, path, and trend.  

```{r, a1_all_1}
a1 %>% 
  tibble::rowid_to_column() %>% 
  ggplot(mapping = aes(x = x, y = y)) +
  geom_path(size = 4.5, color = 'black', alpha = 0.25) +
  geom_line(size = 1.25, color = 'red', linetype = 'dashed') +
  geom_point(size = 10,
             mapping = aes(fill = rowid),
             shape = 21, color = 'black', stroke = 1.) +
  geom_text(mapping = aes(label = rowid,
                          color = rowid > 6)) +
  geom_smooth(formula = y ~ x,
              method = lm,
              se = FALSE,
              size = 1.55) +
  scale_fill_viridis_c(guide = FALSE) +
  scale_color_manual(guide = FALSE,
                     values = c("TRUE" = 'black',
                                "FALSE" = 'white')) +
  theme_bw()
```

