---
title: "CMPINF 2130 Summer 2021 - Week 05"
author: "Dr. Joseph P. Yurko"
date: "6/14/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages

```{r, load_tidyverse}
library(tidyverse)
```

## Review bar chart

Use the `mpg` data set for counts instead of `iris`.  

```{r, glimpse_mpg}
mpg %>% glimpse()
```

The number of observations for each unique value of `class` in `mpg`.  

```{r, class_bar_chart}
mpg %>% 
  ggplot(mapping = aes(x = class)) +
  geom_bar() +
  theme_bw()
```

Reorder the x-axis.  

The character gets converted to a factor variable behind the scenes. A factor is R's terminology for a categorical.  

```{r, check_class_factor}
mpg %>% 
  mutate(class = as.factor(class)) %>%
  select(class)
```


Let's print the factor to the screen.  

```{r, check_class_factor_2}
mpg %>% 
  mutate(class = as.factor(class)) %>% 
  select(class) %>% 
  pull()
```


We can use the `levels()` function to extract the finite set of levels or values the factor can take.  

```{r, check_class_factor_levels}
mpg %>% 
  mutate(class = as.factor(class)) %>% 
  select(class) %>% 
  pull() %>% 
  levels()
```

We can manually override the order by specyfing the levels.  

```{r, relevel_the_class_variable}
mpg %>% 
  mutate(class = factor(class, levels = c("suv", "subcompact", "pickup",
                                          "2seater",
                                          "minivan", "midsize", 'compact'))) %>% 
  select(class) %>% 
  pull() %>% 
  levels()
```


The level order dictates the order of the x-axis in the bar chart.  

```{r, our_new_class_level_order_bar}
mpg %>% 
  mutate(class = factor(class, levels = c("suv", "subcompact", "pickup",
                                          "2seater",
                                          "minivan", "midsize", 'compact'))) %>% 
  ggplot(mapping = aes(x = class)) +
  geom_bar() +
  theme_bw()
```

What if we want to order by count?  

Let's first check by chaining `count()` and `arrange()`.  

```{r, check_class_count_order}
mpg %>% 
  count(class) %>% 
  arrange(n)
```

Extract out the `class` column from the summarized data set.  

```{r, grab_the_class_orders}
class_ascend_order <- mpg %>% 
  count(class) %>% 
  arrange(n) %>% 
  select(class) %>% 
  pull()
```

Now we can order the x-axis based on ascending counts.  

```{r, mpg_class_bar_ascend}
mpg %>% 
  mutate(class = factor(class, levels = class_ascend_order)) %>% 
  ggplot(mapping = aes(x = class)) +
  geom_bar() +
  theme_bw()
```

[forcats](https://forcats.tidyverse.org/) streamlines manipulating categorical variables.  

```{r, check_the_count_order_forcats}
mpg %>% 
  mutate(class = forcats::fct_infreq(class)) %>% 
  select(class) %>% 
  pull() %>% 
  levels()
```


Visualize the counts as a bar chart using the forcats ordering.  

```{r, class_bar_forcats}
mpg %>% 
  mutate(class = forcats::fct_infreq(class)) %>% 
  ggplot(mapping = aes(x = class)) +
  geom_bar() +
  theme_bw()
```

We can reverse the order showing lowest count to highest. Manually call the reverse order function.  

```{r, class_bar_forcats_2}
mpg %>% 
  mutate(class = forcats::fct_infreq(class)) %>% 
  mutate(class = forcats::fct_rev(class)) %>% 
  ggplot(mapping = aes(x = class)) +
  geom_bar() +
  theme_bw()
```

We can also include the count as text.  

```{r, class_bar_forcats_3}
mpg %>% 
  mutate(class = forcats::fct_infreq(class)) %>% 
  mutate(class = forcats::fct_rev(class)) %>% 
  ggplot(mapping = aes(x = class)) +
  geom_bar() +
  geom_text(stat = 'count',
            mapping = aes(label = stat(count)),
            color = 'red', nudge_y = 2, size = 5.5) +
  theme_bw()
```

### Lumping

forcats has a dedicated set of functions to perform the "lumping" for us.  

```{r, lump_class_var}
mpg %>% 
  mutate(class = forcats::fct_lump_min(class, 15)) %>% 
  count(class)
```

Let's now look at a bar chart which includes the lumped "Other" variable.  

```{r, lump_class_bar}
mpg %>% 
  mutate(class = forcats::fct_lump_min(class, 15)) %>% 
  ggplot(mapping = aes(x = class)) +
  geom_bar() +
  theme_bw()
```

Let's reorder by the count.  

```{r, lump_class_bar_2}
mpg %>% 
  mutate(class = forcats::fct_lump_min(class, 15)) %>% 
  mutate(class = forcats::fct_infreq(class)) %>% 
  ggplot(mapping = aes(x = class)) +
  geom_bar() +
  theme_bw()
```

We can rename the "Other" level.  

```{r, lump_class_bar_3}
mpg %>% 
  mutate(class = forcats::fct_lump_min(class, 15, other_level = "The other level")) %>% 
  mutate(class = forcats::fct_infreq(class)) %>% 
  ggplot(mapping = aes(x = class)) +
  geom_bar() +
  theme_bw()
```

Lumping based on less than 15 observations does not scale to larger or smaller data sets. Instead we should consider the **proportion**.  

Let's manually calculate the proportion of each value.  

```{r, mpg_class_prop}
mpg %>% 
  count(class) %>% 
  mutate(prop_value = n / sum(n)) %>% 
  arrange( desc(prop_value) )
```

Check the proportions.  

```{r, check_prop_value}
62 / nrow(mpg)
```


Let's now visualize the levels as proportions instead of counts. Let's first use `geom_col()`.  

```{r, mpg_class_prop_col}
mpg %>% 
  count(class) %>% 
  mutate(prop_value = n / sum(n)) %>% 
  arrange( desc(prop_value) ) %>% 
  ggplot(mapping = aes(x = class, y = prop_value)) +
  geom_col() +
  theme_bw()
```

Reorder the levels.  

```{r, mpg_class_prop_col_2}
mpg %>% 
  count(class) %>% 
  mutate(prop_value = n / sum(n)) %>% 
  arrange( desc(prop_value) ) %>% 
  mutate(class = forcats::fct_inorder(class)) %>% 
  ggplot(mapping = aes(x = class, y = prop_value)) +
  geom_col() +
  theme_bw()
```

If you want the percent, just multiply by 100.  

```{r, mpg_class_prop_col_3}
mpg %>% 
  count(class) %>% 
  mutate(prop_value = n / sum(n)) %>% 
  arrange( desc(prop_value) ) %>% 
  mutate(class = forcats::fct_inorder(class)) %>% 
  ggplot(mapping = aes(x = class, y = 100 * prop_value)) +
  geom_col() +
  theme_bw()
```

If we multiply outside of ggplot.  

```{r, mpg_class_prop_col_4}
mpg %>% 
  count(class) %>% 
  mutate(prop_value = n / sum(n)) %>% 
  arrange( desc(prop_value) ) %>% 
  mutate(class = forcats::fct_inorder(class)) %>% 
  mutate(prop_value = 100 * prop_value) %>% 
  ggplot(mapping = aes(x = class, y = prop_value)) +
  geom_col() +
  # manually override the axis labels
  labs(y = "percent") +
  theme_bw()
```

Ideally, we want ggplot2 to show us the proportions from the original or raw data set. We have to make use of the `stat()` function again.  

```{r, mpg_class_prop_viz}
mpg %>% 
  ggplot(mapping = aes(x = class)) +
  geom_bar(mapping = aes(y = stat(prop))) +
  theme_bw()
```

To override the grouping and visualize the proportion relative to the WHOLE data set we must specify `group = 1` in aes().  

```{r, mpg_class_prop_correct}
mpg %>% 
  ggplot(mapping = aes(x = class)) +
  geom_bar(mapping = aes(y = stat(prop),
                         group = 1)) +
  theme_bw()
```

We can reorder the level as needed.  

```{r, mpg_class_prop_correct_2}
mpg %>% 
  mutate(class = forcats::fct_infreq(class)) %>% 
  ggplot(mapping = aes(x = class)) +
  geom_bar(mapping = aes(y = stat(prop),
                         group = 1)) +
  theme_bw()
```

We can also display the proportion as text.  

```{r, mpg_class_prop_with_text}
mpg %>% 
  mutate(class = forcats::fct_infreq(class)) %>% 
  ggplot(mapping = aes(x = class)) +
  geom_bar(mapping = aes(y = stat(prop),
                         group = 1)) +
  geom_text(stat = 'count',
            mapping = aes(y = after_stat( count / sum(count)),
                          label = after_stat( signif(count / sum(count), 3) )),
            color = 'red', nudge_y = 0.0075, size = 3.5) +
  theme_bw()
```

Lump all values or levels of `class` together if they occur less than 5% of the time. Or, the level **proportion** is less than 0.05.  

```{r, mpg_class_prop_text_lump}
mpg %>% 
  mutate(class = forcats::fct_lump_prop(class, 0.05)) %>% 
  mutate(class = forcats::fct_infreq(class)) %>% 
  ggplot(mapping = aes(x = class)) +
  geom_bar(mapping = aes(y = stat(prop),
                         group = 1)) +
  geom_text(stat = 'count',
            mapping = aes(y = after_stat( count / sum(count)),
                          label = after_stat( signif(count / sum(count), 3) )),
            color = 'red', nudge_y = 0.0075, size = 3.5) +
  theme_bw()
```



## Combinations

`drv` tells us if the vehicle is 4wheel, front, or rear wheel drive.  

```{r, count_per_drv}
mpg %>% count(drv)
```

Let's look at the counts of the combinations of `drv` and `class`.  

```{r, mpg_class_drv_bar}
mpg %>% 
  ggplot(mapping = aes(x = class)) +
  geom_bar(mapping = aes(fill = drv)) +
  ggthemes::scale_fill_calc() +
  theme_bw()
```

When we want to know the proportion **within** the x-aesthetic group we only need to specify the position as `'fill'` to know the proportion.  

```{r, mpg_class_drv_bar_prop}
mpg %>% 
  ggplot(mapping = aes(x = class)) +
  geom_bar(mapping = aes(fill = drv),
           position = 'fill') +
  ggthemes::scale_fill_calc() +
  theme_bw()
```


Use `labs()` to change the y-axis label.  

```{r, mpg_class_drv_bar_prop_2}
mpg %>% 
  ggplot(mapping = aes(x = class)) +
  geom_bar(mapping = aes(fill = drv),
           position = 'fill') +
  ggthemes::scale_fill_calc() +
  labs(y = 'proportion') +
  theme_bw()
```

Add a title.  

```{r, mpg_class_drv_bar_prop_3}
mpg %>% 
  ggplot(mapping = aes(x = class)) +
  geom_bar(mapping = aes(fill = drv),
           position = 'fill') +
  ggthemes::scale_fill_calc() +
  labs(y = 'proportion',
       title = 'Proportion of drv within each vehicle class') +
  theme_bw()
```

Instead, we might be interested in the proportio of the combination directly.  

```{r, mpg_class_drv_combo_prop}
mpg %>% 
  ggplot(mapping = aes(x = class)) +
  geom_bar(mapping = aes(fill = drv,
                         y = after_stat( count / sum(count) ))) +
  ggthemes::scale_fill_calc() +
  theme_bw()
```

Lastly, we can 'dodge' to help see the individual combination proportions.  

```{r, mpg_class_drv_combo_prop_dodge}
mpg %>% 
  ggplot(mapping = aes(x = class)) +
  geom_bar(mapping = aes(fill = drv,
                         y = after_stat( count / sum(count) )),
           position = 'dodge') +
  ggthemes::scale_fill_calc() +
  theme_bw()
```

We can check the counts/proportions.  

```{r, check_the_class_drv_combos}
mpg %>% 
  count(class, drv) %>% 
  mutate(prop_value = n / sum(n))
```

## Continuous variables

We will start with synthetic data. We will generate 5000 random draws or samples from a **standard normal distribution**.  

**standard** refers to the fact that the **mean** is equal to 0 and the standard deviation is equal 1.  

**z-score** has a mean zero and variance one.  

When generating random numbers it's always import to set the random seed.  

```{r, make_fake_data}
set.seed(2130)
x <- rnorm(n = 5000)

x %>% length()
```

Check the summary statistics with `summary()`.  

```{r, check_x_summary_stats}
x %>% summary()
```

Let's put `x` inside a data.frame, or really a tibble.  

```{r, make_fake_tibble}
xdf <- tibble::tibble(x = x)

xdf %>% glimpse()
```

As a continuous variable, `x`, does not have a finite set of values. There's essentially an infinite set of possible values.  

```{r, check_number_of_uniques}
xdf %>% 
  summarise(n_distinct(x))
```

We cannot simply use a bar chart for a continuous number!  

```{r, viz_x_as_factor_bar}
xdf %>% 
  slice(1:200) %>% 
  ggplot(mapping = aes(x = as.factor(x))) +
  geom_bar() +
  theme_bw() +
  theme(axis.text.x = element_blank())
```

Let's print the first 25 values to the screen.  

```{r, check_unique_values_x}
xdf %>% 
  count(x) %>% 
  arrange(x) %>% 
  head(25)
```

We can **bin** similar observations together. A histogram is a "bar chart" for continuous variables. A histogram is made by:  

* "breaking" or dividing the *range* into smaller intervals or **bins**  
* count the number of observations in each bin  
* display a bar chart to show the counts in each bin  

Let's see what happens if we use the quantiles to divide the space.  

```{r, viz_x_quant_bins}
xdf %>% 
  mutate(xbins = cut(x, quantile(x), include.lowest = TRUE)) %>% 
  ggplot(mapping = aes(x = xbins)) +
  geom_bar() +
  theme_bw()
```

Usually best to use **equal** or **uniform** bin sizes or bin intervals when visualizing histograms. Let's use the `seq()` function to define evenly or uniform interval iszes in bins.  

```{r, viz_x_hist_bar}
xdf %>% 
  mutate(xbins = cut(x, 
                     seq(min(x), max(x), length.out = 5),
                     include.lowest = TRUE)) %>% 
  ggplot(mapping = aes(x = xbins)) +
  geom_bar() +
  theme_bw()
```

Use more bins to get better "resolution" of the **distribution** of the variable.  

```{r, viz_x_hist_bar_2}
xdf %>% 
  mutate(xbins = cut(x, 
                     seq(min(x), max(x), length.out = 11),
                     include.lowest = TRUE)) %>% 
  ggplot(mapping = aes(x = xbins)) +
  geom_bar() +
  theme_bw()
```

Use even more bins to get further resolution, but we then have fewer observations in each bin.  

```{r, viz_x_hist_bar_3}
xdf %>% 
  mutate(xbins = cut(x, 
                     seq(min(x), max(x), length.out = 21),
                     include.lowest = TRUE)) %>% 
  ggplot(mapping = aes(x = xbins)) +
  geom_bar() +
  theme_bw()
```

We can just use the `geom_histogram()` geom to create the histogram.  

```{r, viz_x_hist}
xdf %>% 
  ggplot(mapping = aes(x = x)) +
  geom_histogram() +
  theme_bw()
```

I tend to use 50 bins for thousands of observations.  

```{r, viz_x_hist_2}
xdf %>% 
  ggplot(mapping = aes(x = x)) +
  geom_histogram(bins = 50) +
  theme_bw()
```

With a few hundred or less I'll use 20 to 25.  

```{r, viz_x_hist_3}
xdf %>% 
  ggplot(mapping = aes(x = x)) +
  geom_histogram(bins = 21) +
  theme_bw()
```

Let's use 2000 bins.   

```{r, viz_x_hist_4}
xdf %>% 
  ggplot(mapping = aes(x = x)) +
  geom_histogram(bins = 2000) +
  theme_bw()
```

A very small number of bins...might not help either.  

```{r, viz_x_hist_5}
xdf %>% 
  ggplot(mapping = aes(x = x)) +
  geom_histogram(bins = 3) +
  theme_bw()
```

We can also decide the number of bins based on the **binwidth**  

```{r, viz_x_hist_6}
xdf %>% 
  ggplot(mapping = aes(x = x)) +
  geom_histogram(binwidth = 0.5) +
  theme_bw()
```

Use the `after_stat()` function to show the proportion per bin.  

```{r, viz_x_hist_prop}
xdf %>% 
  ggplot(mapping = aes(x = x)) +
  geom_histogram(binwidth = 0.5,
                 mapping = aes(y = after_stat( count / sum(count) ))) +
  theme_bw()
```

Histogram is a representation of the **probability density function** (pdf). We can set the y-axis to the **density** of the bins. Density is calculated such that the integral or area under the curve equals 1.  

```{r, viz_x_hist_density}
xdf %>% 
  ggplot(mapping = aes(x = x)) +
  geom_histogram(binwidth = 0.5,
                 mapping = aes(y = stat(density))) +
  theme_bw()
```

Instead, we could also represent the distribution with a **kernel density estimate**.  

```{r, viz_x_density}
xdf %>% 
  ggplot(mapping = aes(x = x)) +
  geom_density(size = 1.1) +
  theme_bw()
```

The **bandwidth** controls how smooth the estimate is. We can control a multiplier to the estimated bandwidth.  

```{r, viz_x_density_2}
xdf %>% 
  ggplot(mapping = aes(x = x)) +
  geom_density(size = 1.1,
               adjust = 0.2) +
  theme_bw()
```

```{r, viz_x_density_3}
xdf %>% 
  ggplot(mapping = aes(x = x)) +
  geom_density(size = 1.1,
               adjust = 2) +
  theme_bw()
```

We can overlay the density with the histogram by adding two layers together.  

```{r, viz_x_density_and_hist}
xdf %>% 
  ggplot(mapping = aes(x = x)) +
  geom_histogram(binwidth = 0.5, alpha = 0.7,
                 fill = 'steelblue') +
  geom_density(size = 2.5) +
  theme_bw()
```

To overlay the two the histogram must be instructed to plot the density.  

```{r, viz_x_density_and_hist_2}
xdf %>% 
  ggplot(mapping = aes(x = x)) +
  geom_histogram(binwidth = 0.5, alpha = 0.7,
                 fill = 'steelblue',
                 mapping = aes(y = stat(density))) +
  geom_density(size = 2.5) +
  theme_bw()
```

Let's go back to what happened with the `quantile()` defined bins. Let's look at the quantile values on the histogram.  

```{r, viz_x_hist_with_ref}
xdf %>% 
  ggplot(mapping = aes(x = x)) +
  geom_histogram(binwidth = 0.5) +
  geom_vline(xintercept = quantile(xdf$x),
             color = 'red', linetype = 'dashed', size = 1.15) +
  theme_bw()
```

Let's look at the `quantile()` result.  

```{r, check_quantile_x}
xdf %>% pull(x) %>% quantile()
```


The sister plot to the histogram or density is the CDF - the cumulative distribution function.  

```{r, viz_x_cdf}
xdf %>% 
  ggplot(mapping = aes(x = x)) +
  stat_ecdf(size = 1.1) +
  theme_bw()
```

Let's roughly calculate the CDF through our manual binning of the `x` values.  

```{r, check_cdf_calc_1}
xdf %>% 
  mutate(xbins = cut(x,
                     breaks = seq(min(x), max(x), length.out = 11),
                     include.lowest = TRUE)) %>% 
  count(xbins)
```

Let's use `geom_col()` to make this bar chart.  

```{r, viz_hist_geom_col}
xdf %>% 
  mutate(xbins = cut(x,
                     breaks = seq(min(x), max(x), length.out = 11),
                     include.lowest = TRUE)) %>% 
  count(xbins) %>% 
  ggplot(mapping = aes(x = xbins, y = n)) +
  geom_col() +
  theme_bw()
```

Calculate the cumulative or rolling sum.  

```{r, check_cdf_calc_2}
xdf %>% 
  mutate(xbins = cut(x,
                     breaks = seq(min(x), max(x), length.out = 11),
                     include.lowest = TRUE)) %>% 
  count(xbins) %>% 
  mutate(rolling_sum = cumsum(n))
```

Visualize our cumulative sum.  

```{r, viz_check_cdf}
xdf %>% 
  mutate(xbins = cut(x,
                     breaks = seq(min(x), max(x), length.out = 11),
                     include.lowest = TRUE)) %>% 
  count(xbins) %>% 
  mutate(rolling_sum = cumsum(n)) %>% 
  ggplot(mapping = aes(x = xbins, y = rolling_sum)) +
  geom_col() +
  theme_bw()
```

We want the rolling or cumulative proportion.  

```{r, viz_check_cdf_prop}
xdf %>% 
  mutate(xbins = cut(x,
                     breaks = seq(min(x), max(x), length.out = 11),
                     include.lowest = TRUE)) %>% 
  count(xbins) %>% 
  mutate(rolling_sum = cumsum(n)) %>% 
  mutate(rolling_prop = rolling_sum / sum(n)) %>% 
  ggplot(mapping = aes(x = xbins, y = rolling_prop)) +
  geom_col() +
  theme_bw()
```

Let's see what happens if we have fewer observations.  

```{r, viz_x_cdf_small}
xdf %>% 
  slice(1:35) %>% 
  ggplot(mapping = aes(x = x)) +
  stat_ecdf(size = 1.1) +
  theme_bw()
```

Make the small data set.  

```{r, make_x_small}
xsmall <- xdf %>% 
  slice(1:35)
```


Look at the CDF for the small data set with specific quantiles called out.  

```{r, viz_x_cdf_small_with_quants}
xsmall %>% 
  ggplot(mapping = aes(x = x)) +
  stat_ecdf(size = 1.25) +
  geom_hline(yintercept = c(0, 0.25, 0.5, 0.75, 1),
             color = 'purple', size = 1.1) +
  geom_vline(xintercept = quantile(xsmall$x),
             color = 'red', linetype = 'dashed', size = 1.2) +
  theme_bw()
```

